---
title: "A latent trajectory analysis in glycemic control and diabetic retinopathy screening coverage 2011-2023: A municipality-level study"
author:
  - name: Rolando Silva-Jorquera
    orcid: 0000-0002-1244-8350
    email: rolando.silva@pucv.cl
    affiliations:
      - name: University College London, Institute for Global Health, London, United Kingdom
  - name: Kasim Allel
    orcid: 0000-0002-2144-7181
    email: kasim.allel1@lshtm.ac.uk
    affiliations:
      - name: University of Oxford, Nuffield Department of Population Health, Oxford, United Kingdom
  - name: Hassan Haghparast-Bidgoli
    orcid: 0000-0001-6365-2944
    email: h.haghparast-bidgoli@ucl.ac.uk
    affiliations:
      - name: University College London, Institute for Global Health, London, United Kingdom
  - name: Paul Nderitu
    orcid: 0000-0002-8571-0751
    email: paul.nderitu@kcl.ac.uk
    affiliations:
      - name: King’s College London, Section of Ophthalmology, Faculty of Life Sciences and Medicine, London, United Kingdom
      
      - name: King’s College London, Department of Ophthalmology, King’s Ophthalmology Research Unit (KORU), London, United Kingdom
  - name: Alasdair Warwick
    orcid: 0000-0002-0800-2890 
    email: alasdair.warwick.19@ucl.ac.uk
    affiliations:
      - name: NIHR Biomedical Research Centre, Moorfields Eye Hospital NHS Foundation Trust and UCL Institute of Ophthalmology, London, United Kingdom
      
      - name: UCL Institute of Cardiovascular Science, University College London, London, United Kingdom
  - name: Claudio Zett
    orcid: 0000-0003-1839-0927
    email: claudio.zett@pucv.cl
    affiliations:
      - name: Pontificia Universidad Católica de Valparaíso, Valparaíso, Chile
format:
  html:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    code_folding: hide
    theme: readable
  pdf:
    number-sections: true
    colorlinks: true
    keeptex: true
    include-in-header: 
      text: |
        \usepackage{booktabs}
        \usepackage{siunitx}
        \newcolumntype{d}{S[
            input-open-uncertainty=,
            input-close-uncertainty=,
            parse-numbers = false,
            table-align-text-pre=false,
            table-align-text-post=false
         ]}
  docx: default
prefer-html: true
date: 'last-modified'
#date-format: '[This version:] MMMM D, YYYY [<br>(First version: February  16, 2023)]'
#abstract: "**Background:** Diabetes management in Chile varies from that in other countries. The Cardiovascular Program (PSCV) aims to control diabetes and retinopathy screening; however, this indicator has not been consistently monitored. Latent class mixture modeling was used to investigate the factors influencing membership in different subpopulations. **Aim**: This study analyzed PSCV data from 2011 to 2023 to identify unique trajectories in diabetes and diabetic retinopathy screening coverage (DMC and DRSC) among municipalities. The characteristics of deprivation were also examined. The analyses were conducted at the ecological level to identify municipalities with coverage differing from the national average or target. **Methods:** This study obtained data from the Statistical and Health Information Department on individuals with T2DM covered by the public health scheme, aged 15 years or above, and eligible for an annual eye examination in all 346 municipalities in Chile between 2011 and 2023. To have stronger analysis we remove the clasified municiplities by size of DM population, and the the lower quintil was removed.A linear Latent Class Mixed Model was used to derive trajectory classes for DRSC and DGCC, with six clusters identified based on clinical plausibility, model fit, and entropy criteria. Logistic regression was used to determine the impact of municipality characteristics on trajectory-class membership. This study focuses on interpreting the shape of clusters from a theoretical perspective. **Results:**  For each of the six models, the fit criteria were evaluated and the proportions in each trajectory class were calculated, along with the estimated intercepts and slopes for each class, with random effects across municipalities. The one-class model tested was a simple linear model with years predicting coverage and cluster membership, with intercepts representing baseline coverage and slopes indicating average annual change. The one-class model for DGC showed an average coverage of 45.10% at baseline with a small but significant negative slope, whereas the one-class model for DRS showed an average coverage of 28.45% with an annual increase of 1.59. Further analysis will focus on the DRSC's class membership, and the two-class model was preferred for DRSC over the one- and three-class models. In 2011, Class 1 (74.32% of municipalities) had an average coverage of 22.88% and experienced a gradual annual increase of 1.41% until 2023, whereas Class 2 (25.68%) had an average of 44.03% in 2011 and experienced a rapid annual increase of 2.07% until 2023. The unadjusted associations of socioeconomic index predictor factors with municipality DRSC trajectory provided an OR estimate of 0.531 (95% CI, 0.387–0.716), suggesting that as deprivation levels rise, the probability of achieving higher coverage (Class 2) increases. **Conclusion:** The DGCC in LAs in Chile appears to follow only a single trajectory, whereas the DRSC follows two slowly growing differing trajectories. Counterintuitively, the trajectories that followed did not appear to be influenced by deprivation. The group of municipalities following a higher DRSC trajectory is potentially worthy of further study to identify whether there are common local policies or practices that may have contributed to their positive deviance from national trends towards an increase in DRSC."
bibliography: references.bib
csl: vancouver-superscript.csl
tbl-cap-location: top
number-sections: true
execute:
  echo: false
  warning: false
  message: false
  cache: false
editor:
  mode: source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
# Use cache = TRUE if you want to speed up compilation

knitr::opts_knit$set(output.format = "html")  # Set to "html" for HTML output


# A function to allow for showing some of the inline code
rinline <- function(code){
  html <- '<code  class="r">``` `r CODE` ```</code>'
  sub("CODE", code, html)
}

```

```{r}
#| eval: true
#| echo: false
#| message: false

library(tidyverse) #dplyr, tibble, ggplot2, readr, tidyr, stringr, forcats, lubridate, purr
library(knitr)
library(kableExtra)
library(ggtext)
library(Cairo)
library(extrafont)
library(hrbrthemes)
library(directlabels)
library(ggrepel)
library(readxl)
library(extrafont)
library(scales)
library(ggsci)
library(lcmm)
library(MixAll)
require(kml)
require(traj)
require(lmerTest)
require(plyr)
require(psych)
require(fpc)
require(mclust)
require(rcompanion)
library(gridExtra)
library(tidyLPA)
library(MASS)
library(broom)
library(skimr)
library(gtExtras)
library(pander)
library(BayesFactor)
library(modelsummary)
library(gt)
library(gtsummary)
library(survival)
library(xtable)
library(skimr)
library(htmltools)

options(scipen=10000000)


```

**ABSTRACT**

**Purpose:** Diabetic retinopathy screening (DRS) and diabetic glycemic control (DGC) are crucial for managing type 2 diabetes; however, access and outcomes vary regionally. This study used latent class mixture modelling (LCMM) on longitudinal data from Chilean municipalities to identify trajectory patterns in type 2 diabetes mellitus management and DRS coverage (DRSC), and to explore their associations with demographics and deprivation, highlighting subpopulations with unique healthcare needs.

**Methods:** We analyzed 2011–2023 data for adults aged ≥15 years with type 2 diabetes from Chile's public health system, focusing on diabetic glycemic control coverage (DGCC) and DRSC. DGCC reflects the proportion of patients with HbA1c \< 7%, and DRSC indicates those undergoing annual retinopathy screening at the municipal level. Data from the Chilean Public Health Platform excluded smaller municipalities. We employed LCMM to identify whether distinct trajectory patterns existed in the DGCC and DRSC. Additionally, we examined the associations between identified trajectories and factors such as deprivation, urbanicity (rural, mixed, urban), and geographic zones (north, central, and south).

**Results:** The analysis included 300 municipalities (2011-2023) and 292 municipalities (2011-2019), examining the DGCC and DRSC. Baseline DGCC was 46.7% (2011-2023) and 45.1% (2011-2019), with annual declines of -1.03% (p \< 0.001) and -0.42% (p \< 0.001), respectively. DRSC increased from 28.5% (2011-2019) to 31.3% (2011-2023), with annual increases of +1.59% (p \< 0.001) and +0.5% (p \< 0.001), respectively. Latent class analysis identified DGCC one-class trajectory up whereas DRSC followed two trajectory classes. Class 1 (lower coverage) had baselines of 24.8% (2011-2023) and 23.05% (2011-2019), with annual changes of +0.4% (p = 0.068) and +1.38% (p \< 0.001), respectively. Class 2 (higher coverage) had baselines of 43.7% (2011-2023) and 43.58% (2011-2019), with annual changes of +0.69% (p = 0.073) and +2.16% (p = 0.002), respectively. Deprivation (lower ISED scores) was linked to the higher DRSC coverage trajectory: OR 0.66 (2011-2023, 95% CI: 0.50, 0.85) and 0.53 (2011-2019, 95% CI: 0.39, 0.73). Rural areas had higher odds for higher coverage, but no significant in the longest period OR 1.79 (2011-2023, 95% CI: 0.99, 3.21) but meaningful for the pre-covid period OR 2.73 (2011-2019, 95% CI: 1.41, 5.30). Urban areas had lower odds: OR 0.39 (2011-2023, 95% CI: 0.17, 0.84) and non-significant of OR 0.47 (2011-2019, 95% CI: 0.19, 1.19). The southern region showed higher odds but not significant for higher coverage: OR 1.02 (2011-2023, 95% CI: 0.42, 2.51) and OR 1.84 (2011-2019, 95% CI: 1.11, 3.08).

**Conclusions:** DGCC in Chilean municipalities followed a single trajectory, whereas DRSC exhibited two distinct trajectories, one with lower coverage and the other with higher coverage. Lower deprivation levels were associated with a higher likelihood of a higher DRSC trajectory. Further exploration of municipalities with higher DRSC coverage may provide insights into the factors contributing to these patterns, potentially highlighting the local policies or practices that could be effective. Future research should investigate the factors influencing DRSC coverage trajectories and evaluate strategies to improve the DGCC and DRSC coverage.

**Keywords**: Latent class trajectories analysis, diabetic retinopathy screening trajectories, diabetes mellitus

# Introduction

```{=html}
<!--

Previous intro 1 - 

The rise in the prevalence of diabetes mellitus (DM), particularly type 2 DM (T2DM),[@idf_2021] has placed strain on health systems, particularly in the management of vision-threatening issues resulting from poor metabolic control. Diabetic retinopathy (DR) is the leading cause of avoidable blindness and vision impairment in working-age adults.[@burton_lancet_2021] The consequences of inadequate diabetes management, coupled with the lack of early DR detection and treatment, not only cause social and emotional distress,[@fenwick_social_2012] but also limit participation in daily activities and job opportunities.[@cooper_psychological_2020] This, in turn, leads to significant economic repercussions[@pdb52_2019] and has a profound impact on the quality of life,[@trento_quality_2013] presenting substantial public health concerns. Therefore, a robust diabetes management system, including regular DR screening (DRS), is crucial to address this pressing issue. However, the effectiveness of diabetes management and efforts to prevent sight loss vary significantly at the local level in Chile[@SILVA-JORQUERA2021; @guerrero-núñez2017] and other countries[@elek2021] and are influenced by factors such as deprivation, geographical location and level of urbanicity, and may also reflect differences in healthcare service policy and provision related to DM.[@guerrero-núñez2017; @elek2021]

In line with the targets and commitments of the 2030 Agenda for Sustainable Development, the World Health Organization has called on health systems to accelerate actions to improve DM detection and quality of care, including achieving good glycemic control (HbA1c \< 7%) in 80% of the diagnosed population.[@who] On the other hand, DR no longer being the leading cause of blindness among working-age adult in England, after a implementation of a systematic organised DRS nationwide screening program[@liew_comparison_2014] This program, has achieved 80% of the people diagnosed with DM annually.[@Scanlon_2017] Overall, in line with the WHO call, health systems should aims to improve the coverage of people achieving a HbA1C \<7% as a 80%, and improve early detection and time treatment of people sight loss threatened by DM.

The Cardiovascular Health Program (CHP) in Chile aims to monitor trends in individuals with type 2 diabetes mellitus (T2DM) and has established guidelines for monitoring glycated hemoglobin (HbA1c) levels and annual retinopathy (DR) screening for individuals aged 15 years or older.[@minsal2017] In 2021, an analysis was conducted to investigate long-term trends in DRS coverage (DRSC).[@SILVA-JORQUERA2021] The analysis revealed significant heterogeneity and considerable variation in longitudinal trends. Such findings raise the question whether there are distinct trajectories within sub-populations of municipalities about the proportion of individuals achieving HbA1c levels of \< 7% and undergoing annual screening for DR among those with T2DM over time. By identifying potentially different pattern trajectories in these indicators, decision-making and resource allocation can be informed to address the burden of T2DM and vision loss within the health system, in line with the World Health Organization's (WHO) call[@who] and existing evidence for improving this coverage.[@Scanlon_2017]

Latent class mixture modeling (LCMM) is an unsupervised machine learning and certainly a data reduction technique used to identify or uncover sub-populations within a larger population based on observed data.[@berlin_introduction_2014] LCMM has been particularly useful in psychiatric and psychological research, where it can be employed to study the heterogeneity of complex phenomena and address research questions related to treatment effects.[@miettunen2016] Furthermore, its applicability extends to a broader range of topics, including epidemiology.[@viner2018] Therefore, such techniques might assist us in addressing this practical problem, which, to the best of our knowledge, has not yet been studied.

Consequently, this study aimed to fill this research gap by determining whether there are specific groups or sub-populations of municipalities with unique trajectory patterns in achieving HbA1c targets of \<7% and DRSC by analysing data at the population level. Additionally, this study sought to investigate the relationship between area-level demographic and deprivation characteristics with these identified trajectories. Although these analyses were conducted at the ecological level, the results may be useful for identifying groups of municipalities that perform better or worse than the national average, providing valuable information for further research.

Previous intro 2 - 

The global rise in the prevalence of diabetes mellitus (DM), particularly type 2 DM (T2DM), has placed a significant strain on health systems worldwide [@idf_2021]. This burden is particularly acute in managing vision-threatening complications arising from poor glycemic control. Diabetic retinopathy (DR), a leading cause of avoidable blindness and vision impairment in working-age adults [@burton_lancet_2021], poses a major public health concern. Inadequate diabetes management, coupled with a lack of early DR detection and treatment, can have devastating consequences beyond vision loss. Studies suggest it can lead to social and emotional distress [@fenwick_social_2012], limit participation in daily activities and job opportunities [@cooper_psychological_2020], and have significant economic repercussions [@pdb52_2019]. Ultimately, these factors can profoundly impact the quality of life for individuals with T2DM [@trento_quality_2013]. Therefore, a robust diabetes management system, including regular DR screening (DRS), is crucial to address this pressing issue.

However, the effectiveness of these strategies varies significantly at the local level in Chile, as evidenced by studies highlighting inequalities in DRS coverage across municipalities [@SILVA-JORQUERA2021; @guerrero-núñez2017]. Similar findings have been reported in other countries [@elek2021]. These inequalities are likely influenced by factors such as deprivation, geographical location and level of urbanicity, and may also reflect differences in healthcare service policy and provision related to DM management [@guerrero-núñez2017; @elek2021].

Given the substantial public health burden of T2DM and its vision-threatening complications, particularly DR, effective management strategies are crucial. While national guidelines in Chile establish protocols for HbA1c monitoring and annual DRS [@minsal2017], prior studies highlight significant inequalities in achieving these targets across municipalities [@SILVA-JORQUERA2021]. These findings suggest potential heterogeneity within the population regarding glycemic control and DRS coverage. Understanding this heterogeneity is vital for developing targeted interventions to address these inequalities and improve overall diabetes care at the local level.

Traditional statistical approaches often assume homogeneity within the population under study. However, in the context of T2DM management, there might be distinct sub-populations with unique trajectories in achieving glycemic control and undergoing DRS. Latent class mixture modeling (LCMM) is an unsupervised machine learning technique that can identify these sub-populations based on observed characteristics [@berlin_introduction_2014; @miettunen2016]. By applying LCMM to longitudinal data on HbA1c levels and DRS coverage across Chilean municipalities, we aim to reveal unique trajectory patterns in achieving glycemic control and undergoing DRS.

Previous intro 3 -

Diabetes Mellitus (DM), a chronic metabolic disorder characterized by elevated blood glucose levels, affected 537 million people globally in 2021, with projections indicating a further increase to 783 million by 2045.[@idf_2021] Among these cases, 96% (95% CI 95.1--96.8) were attributed to type 2 DM (T2DM), making it the primary contributor to the overall DM burden at 95.4% (95% CI 94.9--95.9)[@ong_global_2023] One-third of the individuals living with DM are affected by diabetic retinopathy (DR),[@burton_lancet_2021] a prevalent complication of T2DM. DR is associated with longer DM duration, poorer glycemic control, and high blood pressure.[@yau_global_2012] Furthermore, DR is the leading cause of moderate-to-severe visual impairment, affecting 1.25% of cases, and accounts for 1.07% of blindness cases.[@burton_lancet_2021] DR it is the leading cause of sight loss worldwide, particularly among working-age adults.[@burton_lancet_2021]

The consequences of sight loss resulting from DM extend beyond individual well-being, encompassing social and emotional stress[@fenwick_social_2012], reduced participation in daily activities, and limited employment opportunities[@cooper_psychological_2020]. This significant public health issue not only impacts the quality of life[@trento_quality_2013] but also incurs substantial economic consequences[@pdb52_2019]. Therefore, an efficient DM management system, including regular DR screening (DRS), is essential for addressing this pressing problem.

For optimal DRS, transitioning from opportunistic to systematically organized DRS is crucial. Opportunistic screening relies on individuals seeking routine health care visits based on their personal circumstances. In contrast, systematically organized programs proactively identify at-risk populations for regular DRS, ensuring broad and consistent coverage. Achieving approximately 80% annual coverage, systematic screening, as exemplified by successful implementation in countries such as England and Wales, significantly reduces sight loss.[@Scanlon_2017] In England, diabetic retinopathy has no longer been the leading cause of blindness among working-age individuals since the implementation of a nationwide DRS program[@liew_comparison_2014], marking a strategic shift in the screening approach.

On the other hand, the extent of diabeteic glycemic control coverage (DGCG) and DRS coverage(DRSC) varies significantly at the global district level[@SILVA-JORQUERA2021; @guerrero-núñez2017, @elek2021]. Socioeconomic factors, coupled with local government policies, play a role in these variations[@guerrero-núñez2017; @elek2021], with pronounced disparities in nations marked by significant income inequality, such as Chile[@lynch2004]. Interestingly, Chile, which lacks a formal DRS programme, has an opportunistic rolling-out DRS. Its highest point in DRSC achieved at 36% in the last decade; less than a half from the 80% target.[@SILVA-JORQUERA2021] However, its specialized eye workforce and a growing eye health infrastructure, positioned to this country in an stage with a significant potential to move towards a systematic DRS approach, leveraging its existing resources and infrastructure to enhance DM management and improve outcomes.

In 2021, an analysis using health datasets at the ecological level from 2011 to 2019 highlighted substantial variations in DRSC when municipalities were grouped by 29 administrative healthcare services (e.g., the lowest group at 15% and the highest averaging 50%, a threefold difference)[@SILVA-JORQUERA2021].Unfortunately, Chile experienced a massive disruption due to COVID-19, mirroring global challenges in managing DM and DRS.[@ahmed2021] Cut DRSC by half and left over 500 thousand individuals with inadequate glycemic control with no annual DRS in 2020.[@silva-jorquera2023] However, there are a still insights to explore DGCC and DRSC to inform policy and practice for better desicion making and resource allocation, specially for planning a systematic DRS. The study prompted inquiries into the existence of different local practices and socioeconomic gradients at the municipality level, including geographical territories, potentially driving DRSC and DGCC.

This study examined the trajectory patterns in the DGCC and DRSC across geographic regions and municipalities in Chile between 2011-2023. We explored the potential impact of COVID-19 on trajectory patterns and the relationship between deprivation characteristics.

-->
```

<!-- The Rising Burden of Diabetes and Diabetic Retinopathy-->

The global increase in diabetes mellitus (DM), particularly type 2 DM (T2DM), has placed an immense burden on healthcare systems worldwide.[@idf_2021] This strain is particularly pronounced in the management of vision-threatening complications arising from inadequate diabetic glycemic control (DGC). Diabetic retinopathy (DR) remains a leading cause of avoidable blindness and vision impairment among working-age adults and poses a critical public health challenge.[@burton_lancet_2021] In response, the World Health Organization (WHO) has emphasized the importance of robust diabetes management and care quality, including maintaining glycated hemoglobin (HbA1c) levels below 7%.[@who] This target aligns with the success observed in England, where a national diabetic retinopathy screening (DRS) program has significantly reduced DR as a cause of blindness in adults.[@liew_comparison_2014]

<!--Inequalities in Diabetes Management and Screening Coverage-->

Despite these efforts, many individuals with T2DM still face social and emotional distress,[@fenwick_social_2012] limited daily functioning and job opportunities,[@cooper_psychological_2020] and significant economic consequences of unmanaged DR and diabetes.[@pdb52_2019] In turn, these impacts profoundly affect quality of life.[@trento_quality_2013] However, access to effective diabetes management and DRS varies significantly across local regions, as seen in Chile, where inequalities in DRS coverage (DRSC) exist among municipalities.[@SILVA-JORQUERA2021; @guerrero-núñez2017] Similar disparities have been documented in other countries, often influenced by socioeconomic factors, urbanicity, and geographic location,[@patel2022; @elek2021; @ravindranath2023; @davis2023; @rawlingschidi2024] which affect access to healthcare and policy implementation for diabetes management.[@guerrero-núñez2017; @elek2021]

<!--Understanding Sub-population Heterogeneity in the Context of Inequalities-->

Addressing these inequalities in Chile requires a clear understanding of the variations in DRS coverage across municipalities. Chile’s national guidelines established by the Cardiovascular Health Program call for regular HbA1c monitoring and annual DRS to prevent sight loss.[@minsal2017] However, there are considerable disparities in meeting these targets across different areas.[@SILVA-JORQUERA2021; @guerrero-núñez2017] Identifying subpopulations with unique healthcare needs and understanding how factors such as demographics and socioeconomic status affect their healthcare trajectories could inform targeted interventions. Such efforts may ultimately improve local-level diabetes management and access to DRS.

<!--Latent Class Mixture Modeling: A Novel Approach-->

Traditional statistical approaches may overlook these population-level variations, often assuming a uniform response to diabetes management across the study population. Distinct subpopulations with unique trajectories for achieving adequate DGC and accessing DRS services are likely to exist. Latent class mixture modelling (LCMM) is an unsupervised machine learning technique that can identify these subpopulations based on observed characteristics.[@berlin_introduction_2014; @miettunen2016] However, to our knowledge, no prior studies have applied LCMM to explore subpopulation heterogeneity within the context of T2DM management and DRSC in Chile.

In this study, we applied LCMM to longitudinal data on T2DM management and DRS coverage in Chilean municipalities. Our aim was to reveal unique trajectory patterns in achieving a DGC, defined as an HbA1c target of\< 7% and access to the DRS. Additionally, we investigated how area-level demographics and deprivation characteristics correlate with these identified trajectories. Although conducted at an ecological level, our findings offer a novel approach to identifying municipalities that perform above or below the national average. These insights have the potential to guide future research and inform policy interventions to reduce sight loss and improve equity in diabetes management.

```{=html}
<!-- METHODS: We analyzed municipality-level data from 2011 to 2023, focusing on adults aged ≥ 15 years diagnosed with type 2 diabetes mellitus (T2DM) and covered by the public health scheme. Data were sourced from the Statistical and Health Information Department, which provides T2DM management metrics through the CHP.[@deis] We used two outcome indicators: the proportion of T2DM patients achieving target glycemic control (HbA1c \< 7%), referred to as diabetic glycemic control coverage (DGCC), and those undergoing annual diabetic retinopathy screening (DRS). DGCC represents the percentage of the T2DM population maintaining an HbA1c level below 7%, while DRS coverage (DRSC) indicates the percentage undergoing annual DRS. Numerators for both metrics were based on the population with glycemic control \< 7% for DGCC and those undergoing annual DRS for DRSC, with the denominator equal to the total T2DM population. Municipalities in the lowest population size quintile were excluded due to their limited T2DM population representation. Due to the incomplete availability of sex-specific data, integrated data were used. To test the associations between various socio-demiographics factors and class membership coverage, we considered deprivation, urbanicity, and geographical location.

For deprivation, we utilized the index of socioeconomic development (ISDE) score, developed in 2013 by the Chilean Observatory of Public Health (OCHISAP). This index evaluates local deprivation levels (ranging from 0 to 1, with lower scores indicating greater deprivation) and comprises three components: income, education, and housing and sanitation.[@gattini_comunas_2013]

Urbanicity was determined by the National Statistics Institute, following the methodological framework outlined in the National Policy of Rural Development.[@bergamini2020] Municipalities were categorized based on population density as predominantly rural (more than 50% of the population residing in low-density areas), mixed (25% to 49% in low-density areas), and predominantly urban (less than 25% in low-density areas).

For geographical location, we used the classification by the Environmental Tribunals of Chile, which subdivides the country into three zones: North, Centre, and South. This classification simplifies the geographical location of municipalities into three levels.

Thus, we analyzed the unadjusted and adjusted associations with deprivation (ISDE score), zone or macro-region (North, Centre, South), and urbanicity (classification as Rural, Mixed, or Urban).

## Statistical analyses

We employed a two-stage analysis. First, we used latent class mixture modeling (LCMM) to analyse DRSC and DRSC longitudinal data. LCMM uses probabilistic assignment to latent classes, which represent subgroups or trajectories and allows the identification of unique patterns within a given population[@berlin_introduction_2014]. We delineated trajectory classes for DRSC and DGCC from 2011 to 2023. We gradually expanded from one class cluster into six classes clusters. Models were evaluated according to their goodness-of-fit, including the Akaike Information Criteria (AIC), Bayesian Information Criteria (BIC), sample-adjusted Bayesian Information Criteria (SABIC), Consistent Akaike Information Criterion (CAIC), Approximate Weight of Evidence Criterion (AWE), Bayes Factor (BF), and correct model probability (cmP). We employed diagnostic tests, including entropy, smallest class count and size, minimun Average Latent Class Posterior Probability (ALCPP). The Vuong--Lo--Mendell--Rubin test assessed overall model fit, comparing a model with k-classes to a model with k-1 classes. The rejection of the k-1 class model in favor of the k-class model was based on a p-value threshold of p\<0.05. When evaluating models, we considered clinical plausibility, model fit, and diagnostic criteria.[@wickrama2021] We hypothesized that the intercept and slope represented random effects across the municipalities. We conducted a sensitivity analysis spanning the time between 2011 and 2019, mirroring our initial analysis to evaluate whether excluding the pandemic years (2019-2023) impacted DRSC and DRSC trajectory results. As LCMM deal with missing data no neccesary to have a complete data for each year in each municipality. However we have checked the data completeness over the time by comparing means using Welch's two sample t -test and verify whether data were balanced between selected trajectories.

Second, we examined the factors influencing membership across different trajectories. We performed logistic regression analysis to determine whether municipal-level socioeconomic development (measured by the ISDE), urbanicity, and macro-region were associated with trajectory-class membership. To ensure the validity of the logistic regression assumptions, the ISDE was standardized to meet the model's distribution requirements.

All statistical analyses were implemented using R version 4.3.1 and the "lcmm" package version 2.1.0.

-->
```

# Methods

## **Study design and data sources**

We conducted a retrospective observational study using annually reported municipality-level health data from 2011 to 2023. Although the data consist of repeated cross-sectional observations collected independently each year, they were analyzed longitudinally to capture temporal patterns and trajectories of care coverage across municipalities. To achieve this, we applied latent class mixed models (LCMM), a probabilistic approach that identifies distinct latent trajectories within heterogeneous populations and is well suited for repeated observations at the aggregate level.[@berlin_introduction_2014] This approach provided a solid foundation for assessing coverage patterns over time.

Our analysis focused on adults aged ≥15 years diagnosed with type 2 diabetes mellitus (T2DM) under the Chilean public health system, consistent with the age threshold established by national cardiovascular program guidelines. Data on T2DM care indicators were sourced from the Statistical and Health Information Department via the Chilean Public Health Platform.[@deis] The data were provided in a curated format as summary indicators at the health centre level and subsequently aggregated at the municipality level for analysis. These data sources laid out the basis for constructing outcomes indicators.

We examined two primary outcomes: diabetic glycemic control coverage (DGCC) and diabetic retinopathy screening coverage (DRSC). DGCC was defined as the proportion of T2DM patients achieving HbA1c \<7% among those monitored annually, while DRSC was defined as the proportion of T2DM patients receiving an annual retinopathy screening. These indicators reflect key aspects of quality in diabetes care provision, offering insight into both metabolic control and preventive screening efforts. It is also important to consider the scope and limitations of these indicators as represented in the dataset, since their interpretation depends on how each was operationalized in the data.

For DGCC, patients without HbA1c testing were excluded from the cardiovascular program after three unsuccessful contact attempts and were therefore not included in the dataset. As a result, the dataset includes only individuals with at least one recorded HbA1c measurement within the calendar year, and those without testing were not assumed to have HbA1c \>7%. This ensures that DGCC reflects actual glycemic control rather than assumptions about missing data. The inclusion process for DRSC was structured differently.

For DRSC, all diagnosed patients remained in the cardiovascular program regardless of whether they attended their annual diabetic retinopathy screening. Hence, all patients were retained in the denominator for this indicator. These inclusion criteria were established by the health system prior to data aggregation.

To improve representativeness, we excluded municipalities in the lowest population-size quintile. These smaller municipalities often exhibited extreme coverage percentages (e.g., 0% or 100%) due to the limited size of their T2DM populations, which could have disproportionately influenced overall estimates and introduced bias into the trajectory modeling. Excluding these municipalities minimized the risk of bias in trajectory estimation and provided a clearer understanding of the relationship between coverage trajectories and key municipal-level characteristics.

We also examined associations between coverage trajectories and key municipal-level characteristics: deprivation, urbanicity, and geographical location. Deprivation was measured using the standardized Index of Socioeconomic Development (ISED), which ranges from 0 (most deprived) to 1 (least deprived) and incorporates indicators such as income, education, housing, and sanitation.[@gattini_comunas_2013] Urbanicity classification followed national guidelines, categorizing municipalities as predominantly rural, mixed, or predominantly urban.[@bergamini2020] Geographical zones were defined using the Environmental Tribunals’ classification, which divides the country into northern, central, and southern zones. These variables were incorporated into the second stage of the analysis

## **Statistical analysis**

We conducted a two-stage analysis to identify coverage trajectory patterns and explore their sociodemographic determinants. In the first stage, we applied latent class mixed models (LCMM) to identify trajectory classes for DRSC and DGCC from 2011 to 2023. LCMM probabilistically assigns latent classes that represent distinct coverage patterns within the population.[@berlin_introduction_2014] In the second stage, we examined whether key municipal-level characteristics were associated with trajectory class membership, following a framework based on established modelling approaches.[@lennon2018] This structure allowed us to both classify trajectory profiles and assess the contextual factors potentially shaping them.

The modelling process began with the construction of a linear fixed-effects scoping model, estimating one to seven latent classes to explore plausible trajectory structures in the absence of prior literature to guide assumptions about class number. Residual patterns from these preliminary models were then used to inform the specification of random effects: flat residuals suggested a random intercept, linear trends indicated a random slope, parabolic curves suggested a quadratic term, and more complex curvatures a cubic term.[@molenberghs_model_2009] Our preliminary models, showed a quadratic o cubic term would be fit well. This step provided a foundation for refining the optimal model based on fit criteria.

After determining the optimal number of classes based on the lowest Bayesian Information Criterion (BIC), we refined the model by testing alternative structures. These included homoscedastic and heteroscedastic models, as well as models with random intercepts or higher-order random effects (see Supplementary Table S1). Model adequacy was assessed using standard criteria: average posterior probability of assignment (APPA \>0.70), odds of correct classification (OCC \>5), relative entropy (\>0.6), mismatch values close to 0, and a minimum class size of 2% of the sample. The 2% threshold was set to ensure meaningful class interpretation in a moderately sized population. If the criteria were not met, the number of classes and model structure were adjusted to enhance classification quality and model fit, and we considered an alternative model with a higher BIC value. We determined the optimal model structure by selecting the two with the lowest BIC value that met the required criteria from the model adequacy assessments thresholds, and compared them to determine which specification offered the most appropriate latent structure. Once we finalized the number of classes and model structure selection and ensured its adequacy, we proceeded to visualize the model outcomes in the different model structures (See Supplementary Material S1).

We used three graphical presentation approaches to visualize model outcomes. The conventional approach involved plotting mean trajectories with time for each class. Additionally, we used mean trajectory plots with 95% predictive intervals for each class, which display the predicted random variation within each class. Another approach was to plot individual-level "spaghetti plots" with time (e.g., using a random sample of participants), allowing the reader to observe changes within classes over time. These three graphical techniques helped provide a comprehensive understanding of the class-specific trajectories. We have also used additional tools for model discrimination.

We use degrees of separation (DoSk) to discriminate the separation of the mean trajectories across latent classes. We applied multivariate Mahalanobis distance to help identifying heterogeneous latent trajectories. Thus, higher DoSK values indicate well-separated mean trajectories, while a DoSK of zero indicates identical trajectories suggesting selection for alternative models with fewer classes.

We use an additional tool to asses the structure of our model trough Elsensohn’s envelope of residualsTo assess the adequacy of model structure assumptions, we implemented a residual-based diagnostic approach originally proposed by Elsenhohn et al. for fixed-effects latent class models and subsequently extended by Lennon et al. to accommodate random effects specifications. For each time point and class, we computed the weighted mean of the predicted coverage values using individuals’ posterior probabilities of class membership as weights. In parallel, we estimated the class-specific and time-specific weighted standard deviation of subject-specific residuals, centered around the weighted mean of residuals. These standard deviations were used to construct upper and lower envelopes (+- 1 SD) around the mean trajectories as a basis for graphical representation of model fit and class-specific uncertainty.

This graphical tool provides a visual assessment of the model’s structure adequacy. Non-parallel boundaries may indicate heteroscedasticity of residuals, while markedly varying interval widths may suggest that within-class heterogeneity is not fully captured. This method is suitable for both fixed- and random-effects latent class trajectory models and contributes to identifying -or confirming - the optimal model structure.

After identifying the optimal model structure, we proceeded with assigning descriptive labels to the latent trajectory classes. Each class was categorized based on observed coverage patterns. This process involved reviewing class-specific characteristics to ensure that each label was meaningful and aligned with plausible real-world patterns. These labels provided a framework for understanding the different patterns of diabetes coverage provision, guiding further interpretation and analysis. Following this, we assessed the robustness of the model by conducting sensitivity analyses.

To assess the robustness of the latent class trajectory classifications, we conducted sensitivity analyses by replicating the modeling on a restricted sample of individuals with at least three observed time points. This threshold ensured sufficient longitudinal information per participant to support reliable class assignment. Agreement between classifications in the full sample and the restricted sample will be quantified using Cohen’s kappa coefficient, providing an objective measure of stability. This diagnostic step allows us to evaluate whether the favored model remains consistent across subsamples, thereby strengthening confidence in its validity and generalizability. After this step, we will assess the interpretability of the favored model by examining potential influential factors.

In the second stage, we examined whether municipality-level characteristics (deprivation, urbanicity, and geographical location) predicted class membership using lineal regression model. The ISED score was standardized to meet distributional assumptions, and both unadjusted and adjusted models were fitted to evaluate the strength and consistency of these associations.

All analyses were conducted in R version 4.4.2, with latent class trajectory modelling performed using the `lcmm` package (version 2.1.0). Additional data processing and visualization tasks were supported by other R packages as appropriate.

This study adhered to the Guidelines for Reporting on Latent Trajectory Studies (GRoLTS) to ensure transparency and replicability in reporting [@vandeschoot2016].

### **Ethics, Funding and Public Involvement**

This study was based on publicly available, aggregate municipality-level data without individual identifiers; therefore, ethical approval was not required. Patients or members of the public were not involved in the design, conduct, or analysis of this study. We plan to disseminate the findings through academic conferences and by engaging with public policy stakeholders and health authorities to support evidence-informed decision-making.

```{=html}
```

# Results

## Overview of DGCC and DRSC Latent Class Trajectory Estimation

### **Latent class number models**

To identify distinct coverage patterns (trajectories) for diabetic glycemic control (DGCC) and retinopathy screening (DRSC), we employed latent class models across two timeframes: 2011-2023 and 2011-2019. This dual-period approach aimed to assess the stability of trajectory structures, especially given the potential post-2019 disruptions. Using the Bayesian Information Criterion (BIC), we identified the following optimal class models in a a scoping model with a homocedastic non random effects model: For DGCC, a seven-class solution was found in both periods, with BIC values of -8434.221 (2011-2023) and -6309.924 (2011-2019). For DRSC, a six-class solution was optimal in both periods, with BIC values of -1997.023 (2011-2023) and -1635.550 (2011-2019). These initial homoscedastic fixed models establish the foundation for evaluating the model structure before testing more complex specifications (e.g., random effects and heteroscedasticity).

### Model structure refinement

To determine the optimal structure for diabetic retinopathy screening coverage (DRSC) and diabetic glycemic control coverage (DGCC), models were analyzed across two time periods (2011-2023 and 2011-2019) while maintaining a fixed number of classes. A seven-class model was tested for both the full period (2011–2023) and the truncated baseline period (2011–2019) for DGCC, whereas a six-class model was used for DRSC. Models were selected based on the lowest Bayesian Information Criterion (BIC). Specifically, for DGCC, the seven-class cubic random effects model (2011-2023) exhibited the lowest BIC of -9137.468, while the seven-class quadratic random effects model (2011-2019) achieved a BIC of -6699.831. For DRSC, the seven-class cubic random effects model (2011-2023) yielded the lowest BIC of -2144.137, and the six-class linear homoscedastic model (2011-2019) had a BIC of -1635.550. These BIC values suggest that the cubic and quadratic models effectively captured complex, nonlinear trajectories for DGCC and DRSC during the full observation period (2011-2023), whereas the linear homoscedastic model indicated a simpler, more stable pattern for DRSC in the truncated period (2011-2019).

### **Model adequacy assessment**

#### DGCC Models (2011–2023)

Two lowest two BIC non-random effect cubic models with three and four latent classes, respectively, met all adequacy thresholds. The four-class model had the lowest BIC (−8,771.2), while the three-class model showed a slightly higher BIC (−8,663.7), with an absolute ΔBIC of 107.5, equivalent to a 1.23% relative increase. This difference represents a **substantially better fit** in favor of the four-class model. The additional (fourth) class remained above the minimum size threshold (4.7%) and retained strong adequacy (APPA = 0.898, OCC = 8.12, entropy = 0.836). Although the three-class solution showed slightly higher classification certainty (APPA = 0.922, entropy = 0.847) and more balanced proportions, both models exhibited similar adequacy overall. Given the significantly lower BIC and comparable classification performance, we favored the four-class model for subsequent interpretation.

#### DGCC Models (2011–2019)

In this restricted period, the two-class cubic random effects model outperformed the alternative seven-class specification both in terms of fit and adequacy. The two-class model had the lowest BIC (−6,708.4), while the seven-class model presented a substantially higher BIC (−6,494.8), with an absolute ΔBIC of 213.6 and a relative increase of 3.18%. This indicates a **significantly better fit** for the two-class model despite its simpler structure. In addition, the two-class solution showed higher entropy (0.887) and posterior probability (0.887), and both classes remained well above the minimum size threshold (8.7%). In contrast, the seven-class model exhibited a decline in classification quality (entropy = 0.801, APPA = 0.783) and a critically small class (2.08%). Given the markedly lower BIC, clearer class separation, and more robust adequacy, the two-class model was retained as the single preferred solution for this period.

#### DRSC Models (2011–2023)

For the DRSC dataset covering the full period, both the four-class cubic non-random effects model and the seven-class quadratic model met adequacy requirements. The four-class model demonstrated a better fit (BIC = −2,076.2), while the seven-class model had a higher BIC (−2,008.5), resulting in an absolute ΔBIC of 67.7 and a relative increase of 3.26%. In addition to this substantially better fit, the four-class model showed stronger classification performance (APPA = 0.854, entropy = 0.791) and maintained all classes above the minimum size threshold. In contrast, the seven-class model presented weaker classification certainty (APPA = 0.752, entropy = 0.759) and introduced several smaller, less stable classes. Given the superior balance between fit, parsimony, and classification adequacy, the four-class model was retained as the preferred solution for this period.

#### DRSC Models (2011–2019)

Within the DRSC models estimated for the 2011–2019 period, two four-class quadratic random effects models were evaluated: one with unrestricted variance and one with a proportional variance structure. The proportional model showed a slightly better fit (BIC = −1,677.4) compared to the unrestricted variant (BIC = −1,675.2), with an absolute ΔBIC of 2.1 and a relative improvement of 0.13%. While the BIC difference was modest, the proportional model also exhibited better classification adequacy, with higher entropy (0.667 vs. 0.635) and posterior probability (APPA = 0.773 vs. 0.749), while maintaining all class sizes comfortably above the minimum threshold (smallest class = 10.8%). Given its marginally superior fit and clearer class separation, the proportional specification was retained as the preferred solution for this period.

Across all models estimated, one final specification was selected for each time period, based on a combination of statistical fit, classification adequacy, and interpretability. For DGCC 2011–2023, the selected model was the four-class cubic non-random effects model, which showed the lowest BIC (−8,771.2) and robust adequacy. For DGCC 2011–2019, the two-class cubic random effects model was retained due to its substantially better fit (ΔBIC = 213.6; 3.18% relative improvement) and high classification certainty. For DRSC 2011–2023, the four-class cubic non-random effects model was preferred for its favorable trade-off between fit (BIC = −2,076.2), entropy (0.791), and class balance. Finally, for DRSC 2011–2019, the four-class quadratic random effects model with proportional variance was selected, showing slightly better fit (ΔBIC = 2.1) and stronger adequacy indicators. These four models constitute the empirical foundation for the scenario-based simulations and comparative analysis presented in the next stage.

### Graphical presentation

Graphical representations were produced for the four selected models, one per dataset and period, illustrating latent class trajectories over time (Figure 1). These include the two-class cubic random effects model (*DGCC 2011–2019*), the four-class cubic non-random effects model (*DGCC 2011–2023*), the four-class quadratic random effects model with proportional variance (*DRSC 2011–2019*), and the four-class cubic non-random effects model (*DRSC 2011–2023*). Each plot displays the estimated mean trajectories by class and highlights differences in growth patterns and timing across latent subgroups.

To complement these plots, individual-level spaghetti graphs were also generated for each model, visualising the distribution and dispersion of observed trajectories within classes (Supplementary Figure S3). These visualisations support the interpretability of class-specific patterns and the temporal dynamics captured by each specification.

```{r}

```

```{r}

```

```{r}

# Create a data frame combining the best scoping models for each outcome and period
scoping_models_summary2 <- bind_rows(
  model_adequacy_table %>%
    filter(str_detect(Model, "7class") & str_detect(Model, "dgcc_model_2011_2023")) %>% 
    select(Model, G, BIC) %>%
    arrange(BIC) %>% 
    slice(1) %>% 
    mutate(Outcome = "DGCC", 
           Period = "2011–2023"),

 model_adequacy_table %>%
    filter(str_detect(Model, "7class") & str_detect(Model, "dgcc_model_2011_2019")) %>% 
    select(Model, G, BIC) %>%
    arrange(BIC) %>% 
    slice(1) %>% 
    mutate(Outcome = "DGCC", 
           Period = "2011–2019"),

  model_adequacy_table %>%
    filter(str_detect(Model, "6class") & str_detect(Model, "drsc_model_2011_2023")) %>% 
    select(Model, G, BIC) %>%
    arrange(BIC) %>% 
    slice(1) %>% 
    mutate(Outcome = "DRSC", 
           Period = "2011–2023"),

  model_adequacy_table %>%
    filter(str_detect(Model, "6class") & str_detect(Model, "class_linear_nre_homocedastic_drsc_model_2011_2019")) %>%
    select(Model, G, BIC) %>%
    arrange(BIC) %>%
    slice(1) %>%
    mutate(Outcome = "DRSC", Period = "2011–2019")
) %>%
  select(Outcome, Period, Model, G, BIC)

# View the resulting table
scoping_models_summary2 

```

```{r}
library(dplyr)
library(stringr)

# Create a data frame combining the best scoping models for each outcome and period
scoping_models_summary <- bind_rows(
  model_adequacy_table %>%
    filter(str_detect(Model, "class_linear_nre_homocedastic_dgcc_model_2011_2023")) %>%
    select(Model, G, BIC) %>%
    arrange(BIC) %>%
    slice(1) %>%
    mutate(Outcome = "DGCC", Period = "2011–2023"),

  model_adequacy_table %>%
    filter(str_detect(Model, "class_linear_nre_homocedastic_dgcc_model_2011_2019")) %>%
    select(Model, G, BIC) %>%
    arrange(BIC) %>%
    slice(1) %>%
    mutate(Outcome = "DGCC", Period = "2011–2019"),

  model_adequacy_table %>%
    filter(str_detect(Model, "class_linear_nre_homocedastic_drsc_model_2011_2023")) %>%
    select(Model, G, BIC) %>%
    arrange(BIC) %>%
    slice(1) %>%
    mutate(Outcome = "DRSC", Period = "2011–2023"),

  model_adequacy_table %>%
    filter(str_detect(Model, "class_linear_nre_homocedastic_drsc_model_2011_2019")) %>%
    select(Model, G, BIC) %>%
    arrange(BIC) %>%
    slice(1) %>%
    mutate(Outcome = "DRSC", Period = "2011–2019")
) %>%
  select(Outcome, Period, Model, G, BIC)

# View the resulting table
scoping_models_summary 
```

We analyzed 300 municipalities during the 2011-2023 period and 292 municipalities during the 2011-2019 period. The trajectories of the DGCC and DRSC over these periods exhibited notable differences. The summary statistics for both outcomes and sociodemographic characteristics are presented in Tables 1 and 2, respectively. Figure 1 (panels a and b) depicts the temporal evolution of municipalities through line trajectories, with individual data points representing the annual coverage values. The longitudinal trajectories of DGCC and DRSC over the 13-year period exhibit unclear patterns, with DRSC showing greater heterogeneity than DGCC. The overall one-class baseline pattern for DGCC revealed a decline in adequate glycemic control over time, while the trajectories for DRSC displayed a rising trend.

1.  For both of our outcomes DRSC and DGCC our preliminary structures by the standardized pattern plots suggested a cubic pattern in the 2011-2023, and a quadratic or cubic pattern for the 2011-2019 period.
2.  After testing 1-7 classes in cubic models for 2011-2023, we have inclined by our top BIC performers, in this case for 2 or 3 classes for DGCC and DRSC. In the 2011-2019 period in their the top performers BIC for DGCC were 4 class model in a cubic structure and 3 class model for a quadratic one, while 3 for a cubic structure and 2 class for a qudrattditc one in DRSC.
3.  We have checked the structure by two and three model class in the eight models structure and confirmed that quadratic and cubic structure are the most bets BIC performers for DRSC in and DGCC in the 2011-2023. Also for the period 2011-2019 we found 2 and 3 claases model as the best performs for DRSC, and 3 and 4 model class in cubic and cuadratic forms for DGCC, respectively.
4.  The top performer in the first approach for 2 and 3 classes was three class model in a cubic structure for DRSC; and 3 and 4 clas model for DGCC in a qudratic and cubic structure respectively.
    1.  For DRSC in 2011-2023 period, our best BIC if for the two-class in a linear random intercept struicture model in the OCC restricted model, while the unrestricted model althouygh was the two-class cubic model. Even when the lowest OCC in the lñast model 3.152 has a considerable highetst BIC best fit structure a for a two class in a cubic model. ALythought the restricted model has more balanced trajectoiries (52 vs 48%), our cubic two classes model offers better fit structure, better entropy and more than acceptable 70% the minim APPA, so this is or selected model. For the same outcome, in 2011-2019, in the restrcicted OCC approach, we obtained the three class model in a linear fixed effect6s, however, the BIC is far higher in the unrestricted OCC but diusvanteged in the lowest trajectory OCC (3.0). When we look at to others cñlasses step back, looking for a better OCC, we found a less parsimonius and less likelihood measures. So our selection its two model class in a quadratic structure.
        -   For DRSC in 2011-2023 = two-class model cubic structure
        -   For DRSC in 2011-2019 = two-class model quadratic structure
    2.  For DGCC in 2011-2023 period, the unrestricted OCC approach has better perormace in two-class model in the linear random incepcept structure for classification (OCC\>5.170), but no better than the unrestricted OCC approach. which has a higher performace in likelihood based metrics, with an accetable OCC of 4.125 and far higher the restricted model (0.59 vs 0.89), therefore we have in line by the two-model class cubic structure. For the period 2011-2019 the restyricted model shows best fit for four-class model, howvwer we found a better structure (BIC -1689) fit in the unresyricted model but gettin a lower OCC in a three class cuadratic model but mostr parsimonius (simple - with a loww number of parameters). So we have step back to corroborate our results looking for othe class model palusibity and we found that class 2 in quadratic structure offer just two classes, but one additional trajectory derived form the larger one, so althoght has a worse correct classification, this structure offer best BIC a most parsimounus one snd meaningful trajectories, so finally we have in clined by 2 classes in a quadratic structure.
        -   For DGCC in 2011-2023 = two-class model cubic structure
        -   For DGCC in 2011-2019 = two-class model quadratic structure
5.  

When the sample size is small, **prioritize likelihood-optimized models**. The likelihood-based metrics (AIC, BIC, log-likelihood) help ensure that the model is not overfitting the small sample, whereas classification metrics might be more prone to instability and overfitting in such cases. Likelihood-based models are more likely to generalize better when the sample size is limited, reducing the risk of spurious or unreliable classification outcomes.

Model adecquacy for the LCMM for both DGCC and DRSC are summarized in Table 3.

<!-- Baseline one-class from our final random intercept model structure for DGCC coverage was slightly higher in the 2011-2023 period (46.7%) than in the 2011-2019 period (45.1%). However, both periods exhibited a significant decline in annual coverage: a -1.03% change per year (p < 0.001) for 2011-2023 and -0.42% per year (p < 0.001) for 2011-2019. In contrast, DRSC coverage increased from 28.5% in 2011-2019 to 31.3% in 2011-2023. The annual change in DRSC was more pronounced in the shorter period, with an increase of 1.59% per year (p < 0.001), compared with 0.5% per year (p < 0.001) in the longer period.-->

## Latent Class Trajectory Models

The analysis examined temporal trends in Diabetic Retinopathy Screening Coverage (DRSC) for two latent classes over the study period (2011-2023). Posterior classification: class1 class2 N 242.00 58.00 % 80.67 19.33. The results revealed that, at the baseline (2011), individuals in the **Moderate Coverage Group** (Class 1) had a DRSC coverage of 13.1%, while individuals in the **High Coverage Growth Group** (Class 2) had a DRSC coverage of 16.2%. Over time, DRSC coverage increased for both groups, with the **Moderate Coverage Group** showing an annual increase of 8.8% and the **High Coverage Growth Group** showing a more pronounced increase of 21.1% per year. However, the quadratic terms suggest that the rate of increase slowed down as the years progressed, and the cubic terms indicate a slight re-acceleration in the later years of the study. These trends were statistically significant (p \< 0.01 for all terms), indicating a strong relationship between DRSC coverage and time for both groups.

The analysis examined temporal trends in Diabetic Retinopathy Screening Coverage (DRSC) for two latent classes over the study period (2011-2019). Posterior classification: class1 class2 N 247.00 53.00 % 82.33 17.67. The results revealed that, at the baseline (2011), individuals in the **Moderate Coverage Group** (Class 1) had a DRSC coverage of 21.7%, while individuals in the **High Coverage Growth Group** (Class 2) had a DRSC coverage of 29.6%. Over time, DRSC coverage increased for both groups. The **Moderate Coverage Group** showed a modest annual increase of 1.3%, while the **High Coverage Growth Group** experienced a more pronounced annual increase of 8.9%. However, the quadratic term for the **High Coverage Growth Group** indicated that the rate of increase slowed down after 2011, with the quadratic term being negative (-0.7%), while the quadratic term for the **Moderate Coverage Group** was not statistically significant, suggesting that the rate of increase remained relatively steady. These trends were statistically significant (p \< 0.01 for Class 2's quadratic term), indicating a strong relationship between DRSC coverage and time for both groups.

The analysis examined temporal trends in Diabetic Glycemic Control Coverage (DGCC) for two latent classes over the study period (2011-2023). Posterior classification: class1 class2 N 279 21 % 93 7. The results revealed that, at the baseline (2011), individuals in the **Moderate Coverage Growth** class had a DGCC of 35.0%, while individuals in the **Rapid Coverage Growth** class had a DGCC of 24.1%. Over time, DGCC coverage increased for both classes. The **Moderate Coverage Growth** class showed a more moderate annual increase of 6.5%, while the **Rapid Coverage Growth** class experienced a more pronounced annual increase of 19.5%. However, the quadratic term for the **Moderate Coverage Growth** class indicated that the rate of increase slowed over time, while the quadratic term for the **Rapid Coverage Growth** class also suggested a deceleration in the growth rate, with both quadratic terms being statistically significant (p \< 0.01). These trends were statistically significant (p \< 0.01 for all terms), indicating a strong relationship between DGCC coverage and time for both classes.

The analysis examined temporal trends in Diabetic Global Control Coverage (DGCC) for two latent classes over the study period (2011-2019). Posterior classification: class1 class2 N 17.00 272.00 % 5.88 94.12. The results revealed that, at the baseline (2011), individuals in the High Coverage with Gradual Decline group had a DGCC of 52.9%, while individuals in the High Coverage with Slower Decrease group had a DGCC of 39.5%. Over time, DGCC coverage increased for the High Coverage with Gradual Decline group, while the High Coverage with Slower Decrease group showed a decline. The High Coverage with Gradual Decline group experienced a slight annual decrease of 5.5%, with a positive quadratic term indicating that the rate of decline slowed over time. In contrast, the High Coverage with Slower Decrease group saw a more substantial annual decrease of 2.9%, but the negative quadratic term indicated that the decline rate was accelerating. These trends were statistically significant (p \< 0.01 for all terms), indicating a strong relationship between DGCC coverage and time for both groups.

The analysis examined temporal trends in Diabetic Retinopathy Screening Coverage (DRSC) for two latent classes over the study period (2011-2023). The results revealed that, at the baseline (2011), Class 1 individuals had a DRSC coverage of **13.1%**, while Class 2 individuals had a DRSC coverage of **16.2%**.

Over time, DRSC coverage increased for both classes, with Class 1 showing an annual increase of **8.8%** and Class 2 showing a more pronounced increase of **21.1%** per year. However, the quadratic terms suggest that the rate of increase slowed down as the years progressed, and the cubic terms indicate a slight re-acceleration in the later years of the study.

These trends were statistically significant (p \< 0.01 for all terms), indicating a strong relationship between DRSC coverage and time for both classes. The model fit was good, as indicated by the AIC and BIC values.

After evaluating models with between one and seven latent classes, the two-class model was identified as the most suitable for DRSC (Supplementary Material). While models with more latent classes achieved slightly lower Bayesian information criterion (BIC) values, the reduction in BIC became negligible beyond five classes. Moreover, models with additional classes demonstrated substantially lower entropy (indicating reduced information quality) and included rare subclasses that overlapped with those in simpler models. As a result, the two-class model was chosen for its balance between fit and simplicity.

For DGCC, the one-class model was selected as the most parsimonious solution, outperforming all other models for both–2011-2023 and 2011-2019 periods. For the DRSC, the two-class model was chosen as the optimal solution for both periods (Table 3).

These trajectory classes reflect significant differences in both baseline coverage and annual changes over the study period.

## Influential Factors in Diabetic Retinopathy Screening Coverage

We analyzed the effects of various factors on DRSC class membership, and the results are summarized in Table 3. Using a logistic regression model, we explored the associations of deprivation (ISED score), urbanicity, and geographical location with DRSC trajectory-class membership.

### Deprivation

#### DRSC

Our analysis revealed significant associations between deprivation and DRSC class membership in both the periods. Higher deprivation (lower ISED scores) was associated with a decreased likelihood of being in a higher DRSC coverage class. The odds ratios (OR) for this association were:

-   **2011-2023 period**: (

    -   Logit(class_membership \~ index_standardized, data=data_reglog_drsc_2011_2023)

        Data Frame: data_reglog_drsc_2011_2023

        Response Variable: class_membership Predictor Variable 1: index_standardized

        Number of cases (rows) of data: 345 Number of cases retained for analysis: 300

        BASIC ANALYSIS

        -- Estimated Model of class_membership for the Logit of Reference Group Membership

        ```         
                        Estimate    Std Err  z-value  p-value   Lower 95%   Upper 95%
           (Intercept)   -1.5017     0.1557   -9.643    0.000     -1.8069     -1.1964 
           index_standardized -0.5226 0.1641 -3.185 0.001 -0.8442 -0.2010
        ```

        -- Odds Ratios and Confidence Intervals

        ```         
                        Odds Ratio   Lower 95%   Upper 95%
           (Intercept)      0.2228      0.1642      0.3023 
           index_standardized 0.5930 0.4299 0.8179
        ```

        -- Model Fit

        ```         
        Null deviance: 294.612 on 299 degrees of freedom
        ```

        Residual deviance: 283.582 on 298 degrees of freedom

        AIC: 287.5819

-   **2011-2019 period**: OR = 0.60 (95% CI: 0.44, 0.83)

    -   Logit(class_membership \~ index_standardized, data=data_reglog_drsc_2011_2019)

        Data Frame: data_reglog_drsc_2011_2019

        Response Variable: class_membership Predictor Variable 1: index_standardized

        Number of cases (rows) of data: 345 Number of cases retained for analysis: 288

        BASIC ANALYSIS

        -- Estimated Model of class_membership for the Logit of Reference Group Membership

        ```         
                        Estimate    Std Err  z-value  p-value   Lower 95%   Upper 95%
           (Intercept)   -1.3873     0.1524   -9.102    0.000     -1.6861     -1.0886 
           index_standardized -0.5064 0.1627 -3.113 0.002 -0.8252 -0.1876
        ```

        -- Odds Ratios and Confidence Intervals

        ```         
                        Odds Ratio   Lower 95%   Upper 95%
           (Intercept)      0.2497      0.1852      0.3367 
           index_standardized 0.6027 0.4382 0.8290
        ```

        -- Model Fit

        ```         
        Null deviance: 294.762 on 287 degrees of freedom
        ```

        Residual deviance: 284.256 on 286 degrees of freedom

        AIC: 288.2564

        Number of iterations to convergence: 4

    -   

These findings indicate that municipalities with higher deprivation scores (less deprived) are less likely to belong to the higher-coverage DRSC class.

#### DGCC

Our analysis revealed non significant associations between deprivation and DGCC class membership in both the periods. The odds ratios (OR) for this association were:

-   **2011-2023 period**: OR = 0.84 (95% CI: 0.53, 1.33)

    -   Logit(class_membership \~ index_standardized, data=data_reglog_drsc_2011_2023)

        Data Frame: data_reglog_drsc_2011_2023

        Response Variable: class_membership Predictor Variable 1: index_standardized

        Number of cases (rows) of data: 345 Number of cases retained for analysis: 300

        BASIC ANALYSIS

        -- Estimated Model of class_membership for the Logit of Reference Group Membership

        ```         
                            Estimate    Std Err  z-value  p-value   Lower 95%   Upper 95%
               (Intercept)   -2.5980     0.2287  -11.358    0.000     -3.0463     -2.1496 
        index_standardized   -0.1767     0.2353   -0.751    0.453     -0.6378      0.2844 
        ```

        -- Odds Ratios and Confidence Intervals

        ```         
                            Odds Ratio   Lower 95%   Upper 95%
               (Intercept)      0.0744      0.0475      0.1165 
        index_standardized      0.8380      0.5284      1.3290 
        ```

-   **2011-2019 period**: OR = 0.92 (95% CI: 0.56, 1.5)

    -   Logit(class_membership \~ index_standardized, data=data_reglog_drsc_2011_2019)

        Data Frame: data_reglog_drsc_2011_2019

        Response Variable: class_membership Predictor Variable 1: index_standardized

        Number of cases (rows) of data: 345 Number of cases retained for analysis: 288

        BASIC ANALYSIS

        -- Estimated Model of class_membership for the Logit of Reference Group Membership

        ```         
                            Estimate    Std Err  z-value  p-value   Lower 95%   Upper 95%
               (Intercept)    2.7784     0.2514   11.050    0.000      2.2856      3.2712 
        index_standardized   -0.0825     0.2491   -0.331    0.740     -0.5707      0.4057 
        ```

        -- Odds Ratios and Confidence Intervals

        ```         
                          Odds Ratio   Lower 95%   Upper 95%
               (Intercept)     16.0925      9.8311     26.3416 
        index_standardized      0.9208      0.5651      1.5003 
        ```

    -   

These findings indicate that municipalities with higher deprivation scores (less deprived) are less likely to belong to class 2 lower dgcc class, yet not significant.

### Urbanicity

#### DRSC

Urbanicity has a varied impact on DRSC coverage, with rural areas generally showing higher odds of improved coverage than mixed-urban zones.

-   **2011-2023 period**: lowerLower of 2.2 (95% CI: 1.07, 4.5), indicating higher odds of being in the higher coverage class compared to mixed urban areas (reference category). Urban areas had an OR of 0.43 (95% CI: 0.15, 1.22), showing lower odds of belonging to the higher DRSC class compared to mixed urban areas yest not statistically significant.

-   **2011-2019 period**: Rural areas had an OR of 2.97 (95% CI: 1.42, 6.2), indicating significantly higher odds of being in the higher coverage class compared to mixed zones. Urban areas had an OR of 0.56 (95% CI: 0.21, 1.55), suggesting lower odds of being in the higher coverage class relative to the reference group, although this was not significant.

Overall, rural municipalities tended to have higher DRSC coverage, whereas urban municipalities showed more variable results.

#### DGCC

Urbanicity has a varied impact on DRSC coverage, with rural areas generally showing higher odds of improved coverage than mixed-urban zones.

-   **2011-2023 period**: Rural areas had an OR of 0.72 (95% CI: 0.29, 1.79), indicating higher odds of being in the higher coverage class compared to mixed urban areas (reference category). Urban areas had an OR of 0 compared to mixed urban areas yest not statistically significant, this is because the lower number of the smaller sample size in not belonging to urban areas, and therefore not posble computing them. (21 municiplities assigned to the smaller class)

-   **2011-2019 period**: Rural areas had an OR of 2.44 (95% CI: 0.74, 7.99), indicating significantly higher odds of being in the higher coverage class compared to mixed zones. Urban areas had an OR of 2.68 (95% CI: 0.85, 8.47), suggesting was not significant.

-   

### Geographical Location

#### DRSC

Geographical location significantly influenced DRSC class membership, with the central region serving as the reference category because of its representative demographic profile and generally better access to healthcare service provision.[@núñez2021]

-   **2011-2023 period**: The southern region had an OR of 1.74 (95% CI: 0.96, 3.16), compared to the central region (reference category), while the northern region had a relatively neutral impact which had an OR of 0.60 (95% CI: 0.17, 2.12).

-   **2011-2019 period**: The southern region had a higher odds ratio of 1.50 (95% CI: 0.84, 2.7), suggesting a trend to being in the hhigher DRSC coverage yet not significant, while the northern region had a relatively neutral impact which had an OR of 0.32 (95% CI: 0.07, 1.44))

These findings indicate that municipalities in the southern region are more likely to achieve higher DRSC coverage than other regions.

#### DGCC

Geographical location significantly influenced DRSC class membership, with the central region serving as the reference category because of its representative demographic profile and generally better access to healthcare service provision.[@núñez2021]

-   **2011-2023 period**: The southern region had an OR of 1.59 (95% CI: 0.57, 4.4), compared to the central region (reference category), while the northern region had a relatively neutral impact which had an OR of 4.19 (95% CI: 1.23, 14).

-   **2011-2019 period**: The southern region had a higher odds ratio of 2.68 (95% CI: 0.85, 8.47), suggesting a trend to being in the hhigher DRSC coverage yet not significant, while the northern region had a relatively neutral impact which had an OR of 0.32 (95% CI: 0.07, 1.44))

These findings indicate that municipalities in the southern region are more likely to achieve higher DRSC coverage than other regions.

Assessment of structure assumptions using residual-based SD envelopes

Residual SD envelope diagnostics supported the adequacy of the cubic fixed-effects model structure. The class-specific envelopes revealed distinct predicted trajectories, with generally consistent boundaries over time. Wider envelopes for Class 1 at the time extremes suggest increased within-class variability or lower data density, whereas Classes 2 and 4 showed narrower, more stable intervals. These patterns suggest that the fixed-effects structure adequately captures between-class separation and temporal dynamics.

```{=html}
<!--
## Unadjusted Associations

### Deprivation

In both the 2011-2023 and 2011-2019 periods, the unadjusted analysis revealed a significant association between the standardized index and class membership. Higher values of the standardized index were consistently associated with a decreased likelihood of belonging to the higher coverage class (OR = 0.665, 95% CI \[0.506, 0.864\], p = 0.00276 for 2011-2023; OR = 0.665, 95% CI \[0.506, 0.864\], p = 0.00276 for 2011-2019 for 2011-2019).

### Urbanicity

Similarly, classification type showed significant associations in unadjusted models for both periods. Municipalities classified as Rural were consistently more likely to belong to the higher coverage class compared to those classified as Mixed (OR = 2.10, 95% CI \[1.16, 3.88\], p = 0.016 for 2011-2023; OR = 2.15, 95% CI \[1.27, 3.72\], p = 0.005 for 2011-2019). The Urban classification exhibited a negative association with higher coverage membership, although non-significant (OR = 0.47, 95% CI \[0.21, 1.01\], p = 0.058 for 2011-2023; OR = 0.58, 95% CI \[0.32, 1.05\], p = 0.071 for 2011-2019)

### Geographical location

This variable also demonstrated significant associations in the unadjusted analysis for both periods. While being in the North zone showed a positive but non-significant association with higher coverage membership compared to the central zone, individuals in the South zone exhibited a positive association, though not statistically significant (OR = 1.51, 95% CI \[0.904, 2.53\], p = 0.116 for 2011-2023; OR = 1.45, 95% CI \[0.83, 2.51\], p = 0.189 for 2011-2019).

## Multivariate Associations

### Incorporating Multiple Variables

When adjusting for multiple variables, including urbanicity and geographical location, the association found between the socio-ecnonomic development and class membership in the unudjusted model lost significance in both periods (OR = 1.02, 95% CI \[0.706, 1.47\], p = 0.903 for 2011-2023; OR = 1.02, 95% CI \[0.706, 1.47\], p = 0.903 for 2011-2019). This suggests that the predictive power of the deprivation is influenced by the inclusion of other factors.

### Persisting Effects of Classification and Zona

Despite the loss of significance for the standardized index in the adjusted models, the Rural classification retained its significance in predicting higher coverage membership for both periods (OR = 2.18, 95% CI \[1.16, 4.20\], p = 0.0168 for 2011-2023; OR = 2.24, 95% CI \[1.27, 3.99\], p = 0.005 for 2011-2019), indicating its robust influence over time. Additionally, trends towards significance for the Urbana classification and the positive association of the Sur zone with higher coverage membership persisted, although not statistically significant.

### Persisting Effects of Urbanicity and Zona

Despite the loss of significance for the standardized index in the adjusted model, the Rural classification retained its significance, though with a slight attenuation in effect size (OR = 2.18, 95% CI \[1.16, 4.20\], p = 0.0168). Additionally, the classification type continued to show trends towards significance, while the zona variable did not maintain significant associations in the adjusted models.

Furthermore, the additional analysis covering the 2011-2019 period reaffirmed the findings from the extended 2011-2023 period. In the unadjusted analysis, significant associations were observed for predictors of higher coverage membership. Individuals classified as Rural were more likely to belong to the higher coverage class compared to those classified as Mixta (OR = 2.15, 95% CI \[1.27, 3.72\], p = 0.005). The Urbana classification showed a negative association with higher coverage membership, albeit non-significant (OR = 0.58, 95% CI \[0.32, 1.05\], p = 0.071). Regarding the zona variable, individuals in the Sur zone exhibited a positive association with higher coverage membership, although not statistically significant (OR = 1.45, 95% CI \[0.83, 2.51\], p = 0.189).

When adjusting for multiple variables, including classification and zona, the associations observed in the unadjusted analysis persisted. The Rural classification remained significant (OR = 2.24, 95% CI \[1.27, 3.99\], p = 0.005), indicating its robust influence on higher coverage membership. Similarly, the trends towards significance for the Urbana classification and the positive association of the Sur zone with higher coverage membership persisted, although not statistically significant.

-->
```

# Discussions

## Summary of Key Findings

We used unsupervised machine learning to identify different subpopulation trajectory patterns for DGCC and DRSC in Chilean municipalities over 12 years. For both the 2011-2023 and 2011-2019 periods, the trajectory patterns for DRSC revealed two distinct patterns. The first depicted municipalities with stable baseline coverage levels, hovering at 27.85% (SE: 2.51) in the 2011-2023 period and slightly lower at 23.05% (SE=1.65%) in the 2011-2019 period. The second trajectory consistently showed higher baseline coverage levels, at 42.14% in the 2011-2023 period and 43.58% in the 2011-2019 period, with significant annual increases in coverage ranging from 1-2%.

Higher deprivation levels (less deprived) were consistently associated with lower DRSC. Rural areas generally showed better coverage than mixed urban environments, while urban areas displayed mixed results. Geographically, municipalities in the southern region consistently exhibited higher DRSC odds ratios compared to the central region. In the adjusted analysis, the influence of urbanicity emerged as the most salient predictor, with rural areas consistently demonstrating higher odds of better coverage.

## What is already known on this topic

Using the same dataset, we have identified significant variation in linear trends in DRSC between Chilean municipalities over the past decade. On the one hand, while deprivation is a predictor for poor uptake, the use of telemedicine has been shown to improve access to diabetic eye care in underserved communities, which often include deprived areas.[@rawlingschid2024] Deprivation affects not only the ability to afford healthcare but also the availability and utilization of services, as it is often associated with geographic isolation and limited healthcare infrastructure.[@lewandohundt2012] This technique has increased interest in health services for obesity,[@viner2018] food addiction,[@aloi2024] loneliness in adolescence,[@verboon2022] risk patterns of myopia[@li2023], or treatment of substance addiction disorders.[@bórquez2024]

In our study, we found that areas characterized by higher levels of deprivation surprisingly exhibited better DRSC. This unexpected finding challenges conventional assumptions and suggests that despite facing socioeconomic challenges, these areas may have prioritized healthcare resources towards DRS services. Possible explanations could include targeted healthcare interventions, focused on the importance of regular screening in deprived communities. Certainly, the use of telemedicine and the adoption of more than 150 primary eye-care units across the country may play a significant role. Deprivation may still be relevant in understanding DRS coverage, but its independent effect may be better explained by other factors included in the model as urbanicity. In fact, deprivation lost its significance after adjusting for urbanicity. While some studies indicate that rural areas may have lower screening coverage compared to urban areas,[@tóth2019] other initiatives have demonstrated the feasibility and effectiveness of targeted screening programs in rural settings, [@hall2022; @khou2021] suggesting that targeted interventions can overcome barriers to screening coverage.[@hall2022] In other words, while socioeconomic status alone may be associated with higher coverage membership, its effect is mediated or influenced by factors such as rural/urban level.

The trajectories of DGCC and DRSC were observed to differ from one another. We found DGCC to present a single trajectory encompassing all 301 municipalities, whereas DRSC was characterized by two probable trajectories without significant slopes between 2011 and 2023. While the overall trend for DGCC indicated a decreasing adequate glycemic control over time, the DRSC trajectories showed an increasing trend, albeit still far from an ideal target - 80% in coverage- to tackle sight loss to long term as been reported.[@Scanlon_2017] It should be noted that the COVID-19 pandemic may have had a disruptive impact on healthcare services, including diabetes monitoring and screening,[@ahmed2021] and could have influenced trend analyses. Excluding the pandemic period yielded consistent results but with significant changes in trajectories over time for the DRSC. It is important to exercise caution when interpreting longitudinal trends that may have been influenced by pandemic data, as demonstrated by our sensitivity analyses.

In the unadjusted analysis, we observed a significant association between the standardized ISED and class membership. This finding underscores the importance of considering the socioeconomic status, as represented by the standardized index, in predicting membership in the higher coverage class. The standardized index serves as a proxy for various socioeconomic factors that may influence an individual's access to healthcare services, such as income level, education, and employment status.

However, when we adjusted for multiple variables, including classification type and zona, the significance of the association between the standardized index and class membership disappeared. This suggests that the predictive power of the standardized index is confounded by the inclusion of other factors.

For example, individuals classified as Rural may have lower socioeconomic status on average compared to those classified as Urban, which could contribute to the observed association between the standardized index and class membership. However, once we account for the Rural classification in the multivariate analysis, the direct effect of the standardized index becomes less significant. This indicates that much of the variation in higher coverage membership explained by socioeconomic status is already captured by the classification variable.

Therefore, while the standardized index provides valuable insight into the socioeconomic determinants of higher coverage membership, its significance in predicting membership diminishes when other relevant factors are taken into consideration. This highlights the need for comprehensive multivariate analyses to better understand the complex interplay between socioeconomic status and other predictors of healthcare coverage.

Interestingly, higher levels of deprivation did not lead to less favorable trajectories for municipalities. Instead, they were associated with an increased likelihood of a high DRSC coverage. This association is particularly pronounced between the period 2011-2019. Although it remained highly significant in the 2011-2023 timeframe, the OR exhibited a smoother association, robustly supporting our findings. On the other hand, unadjusted association with urbanicity and their three categories showed that the Urban population is the less favourable in achieving the highuer coverage trajectory, meanwhile Rural population the most favourable, though non statistically significant. It seems urbanisation could explain better the association with DRSC trajectories than deprivation as this lost its significant effect when we adjusted by urbanisation level. Therefore, understanding the factors that influence DRSC is crucial. Future research should explore these practical implications. Potentially, factors linked to targeted health policies addressing diabetes within Chile's 2005 health reform[@bastias2008] and financing primary healthcare based on factors such as deprivation through capitation[@cuadrado2022] could provide insights and explain, at least in part, our findings.

To date, there is no evidence of the use of LCMM techniques to investigate geographical differences in the DRSC in Chile or other regions. Similar evidence has explored obesity prevalence in childhood across local authorities in the UK, highlighting its significance in studying geographical data to inform policies.[@viner2018] Despite previous discussions, no studies at the municipal level have used mixed-effect models or LCMM techniques to identify unique trends in the DRSC. LCMM methods enable the identification of municipalities that demonstrate exceptional progress in longitudinal trends for sight-loss-related diabetes prevention, enhance the understanding of geographical inequalities, and inform targeted interventions to improve screening coverage.

Utilizing PSCV data from more than a decade, we successfully identified distinct subpopulations of municipalities that exhibited varying trajectories in the DRSC. Most municipalities displayed a high probability of belonging to their designated classes, suggesting a strong alignment between these assigned classes and DRSC from 2011 to 2023. However, some municipalities exhibited notably lower probabilities of being assigned to the DRSC classes, indicating potential inaccuracies. Although our analyses were conducted at the ecological (municipality-level) rather than at the individual level (T2DM patient), this approach was considered appropriate as a key determinant of healthcare access for DRSC often operating at the primary healthcare level. Our primary objective was to identify the potential subpopulations among municipalities. Nevertheless, we could not establish causality for the relationships identified between socioeconomic deprivation and municipality trajectory classes.

While the trajectory classes we identified are plausible, LCMM is an exploratory, data-driven technique, and chance relationships in our data may have influenced our findings regarding trajectory groups. However, we were cautious in identifying the most likely latent class groups and the trajectories we identified were reasonable. The southern regions exhibited a notable trend, with a higher likelihood of falling into the upper DRSC category, suggesting distinct regional patterns potentially influenced by socioeconomic conditions or healthcare accessibility in these areas. Further investigation into the impact of regional policies on DRSC access could be beneficial, encompassing factors such as the availability of eye care specialists, the implementation of mobile clinics for underserved populations, and targeted initiatives to enhance access in less advantaged regions. These area-specific efforts could play a crucial role in broadening screening coverage and mitigating regional healthcare inequalities.

Our data on municipality deprivation took place in 2013, but we hypothesized that it is possible that the overall deprivation levels remained relatively stable over the thirteen-year (and nine-year) study period. Nonetheless, trajectory class membership may have been influenced by changes in deprivation levels during the study period.

# Conclusions

The existence of DGCC in Chilean municipalities appears to follow a single trajectory, whereas the DRSC presents two distinct, gradually increasing trajectories after testing the pandemic period. Higher deprivation levels (lower ISED score) were associated with an increased likelihood of higher DRSC trajectory. It is worthwhile to examine municipalities that demonstrate a higher DRSC trajectory (see [supplementary data](https://github.com/Rolo-Silva/trajectories_diabetes)) in greater detail to identify any shared local policies or practices that may contribute to their positive deviation from national trends in DRSC increase. Future studies should investigate the factors that influence DRSC coverage classes over time, and consider the implications of enhancing health systems in relation to DGCC/DRSC trajectories.

**Acknowledgements**

**Author contribution.**

**Conflict of interest**

Authors declared they have no conflicts of interest.

**Funding**

This study forms part of the author’s doctoral research, supported by a scholarship from the Chilean National Agency for Research and Development (ANID). Funding to present the results at academic conferences was provided by University College London (UCL). The funders had no role in the design, conduct, analysis, or reporting of the study.

```{=html}
<!--
#| label: fig-1
#| fig-cap: "Raw data coverage trajectories 2011-2023"
#| fig-subcap: 
#| - "Diabetic Glycemic Control Coverage"
#| - "Diabetic Retinopathy Screening Coverage"
#| layout-ncol: 2
#| fig-align: center

# Plots for raw trajectories for DRSC and DGCC --------------------

coverage1 <- read.csv("coverage_2011_2023_noq1.csv", sep=",") #read coverage file
#coverage2 <- read.csv("coverage_2011_2019_noq1.csv", sep=",")


ggplot(coverage1, aes(x=ano2, y=dm_coverage)) +
  geom_point(size = .2) + 
  geom_line(size = .2, aes(group = comuna2)) +
  scale_x_continuous(breaks = c(0:12), labels=2011:2023)+
  xlab("Year") +
  ylab("Proportion of T2DM individuals with HbA1C < 7%")+
  #theme_ipsum()+
 theme(panel.grid.major = element_blank(), # optional - remove gridlines
        panel.grid.minor = element_blank(),
        panel.background = element_blank())+
  theme(legend.position = "none") +
  #geom_smooth(se = T, method = "loess", aes(group = class))+
  scale_color_lancet()


ggplot(coverage1, aes(x=ano2, y=drs_coverage)) +
  geom_point(size = .2) + 
  geom_line(size = .2, aes(group = comuna2)) +
scale_x_continuous(breaks = c(0:12), labels=2011:2023)+
  xlab("Year") +
  ylab("Proportion of T2DM with an annual eye screening")+
  #theme_ipsum()+
  labs(colour= "class magnitude")+
  theme(legend.position = "none") +
  #geom_smooth(se = T, method = "loess", aes(group = class))+
 # scale_color_lancet(labels = c("82.27%", "17.73%")) 
  theme(panel.grid.major = element_blank(), # optional - remove gridlines
        panel.grid.minor = element_blank(),
        panel.background = element_blank())+
  theme(legend.position = "none") +
  scale_color_lancet()

-->
```

```{r}
#| label: fig-1
#| warning: false
#| fig-cap: "Raw and baseline one-class trend data over thirteen-year coverage trajectories for 300 municipalities (2011–2023), with lines connecting dots marking annual coverage"
#| fig-subcap: 
#| - "Diabetic Glycemic Control Coverage"
#| - "Diabetic Retinopathy Screening Coverage"
#| layout-ncol: 2
#| fig-align: center

dm1 <- read.csv("dm_2011_2023_noq1.csv", sep=",") 

ggplot(dm1, aes(x=ano2, y=dm_coverage, colour = as.factor(class), group = comuna2)) +
  geom_point(size = .1) + 
  geom_line(size = .05) +
  scale_x_continuous(breaks = c(0:12), labels=2011:2023)+
  xlab("Year") +
  ylab("Proportion of T2DM individuals with HbA1C < 7%")+
  #theme_ipsum()+
  theme(legend.position = "none") +
  geom_smooth(se = T, method = "loess", aes(group = class))+
   theme(panel.grid.major = element_blank(), # optional - remove gridlines
        panel.grid.minor = element_blank(),
        panel.background = element_blank())+
  scale_color_lancet()

dr1 <- read.csv("drs_2011_2023_noq1.csv", sep=",") 

ggplot(dr1, aes(x=ano2, y=drs_coverage, colour = as.factor(class), group = comuna2)) +
  geom_point(size = .1) + 
  geom_line(size = .05) +
  geom_smooth(method='lm',formula=drs_coverage~ano2)+
  scale_x_continuous(breaks = c(0:12), labels=2011:2023)+
  xlab("Year") +
  ylab("Proportion of T2DM with an annual eye screening")+
  #theme_ipsum()+
  #labs(colour= "class magnitude")+
  theme(legend.position = "none") +
  geom_smooth(se = T, method = "loess", aes(group = class))+
  #guides(color = guide_legend(override.aes = list(fill = NA)),
  #       linetype = guide_legend(override.aes = list(fill = NA))) +
  theme(panel.grid.major = element_blank(), # optional - remove gridlines
        panel.grid.minor = element_blank(),
        panel.background = element_blank())+
  scale_color_lancet()
  

```

```{r}
#| label: fig-2
#| warning: false
#| fig-cap: "Latent class trajectory mixture models solutions for diabetic retinopathy screening coveragein three differente periods; 2011-2023; 2011-2019; and 2020-2023"
#| fig-subcap: 
#| - "2011-2023"
#| - "2011-2019"
#| - "2020-2023"
#| layout-ncol: 1
#| fig-align: center

dr2 <- read.csv("drs2_2011_2023_noq1.csv", sep=",") 

ggplot(dr2, aes(x=ano2, y=drs_coverage, colour = as.factor(class), group = comuna2)) +
  geom_point(size = .1) + 
  geom_line(size = .05) +
  geom_smooth(method='lm',formula=drs_coverage~ano2)+
  scale_x_continuous(breaks = c(0:12), labels=2011:2023)+
  xlab("Year") +
  ylab("Proportion of T2DM with an annual eye screening")+
  #theme_ipsum()+
  #labs(colour= "class magnitude")+
  geom_smooth(se = T, method = "loess", aes(group = class))+
  guides(color = guide_legend(override.aes = list(fill = NA)),
         linetype = guide_legend(override.aes = list(fill = NA))) +
  theme(panel.grid.major = element_blank(), # optional - remove gridlines
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        #legend.box.background = element_rect(colour = "red"),
        legend.key = element_rect(fill = "white", colour = "white"))+
  #scale_color_lancet()+
  scale_color_lancet(labels = c("202 trajectories (67.3%)", "98 trajectories (32.7%)")) 


dr3 <- read.csv("drs2_2011_2019_noq1.csv", sep=",") 

ggplot(dr3, aes(x=ano2, y=drs_coverage, colour = as.factor(class), group = comuna2)) +
  geom_point(size = .1) + 
  geom_line(size = .05) +
  geom_smooth(method='lm',formula=drs_coverage~ano2)+
  scale_x_continuous(breaks = c(0:12), labels=2011:2023)+
  xlab("Year") +
  ylab("Proportion of T2DM with an annual eye screening")+
  #theme_ipsum()+
  #labs(colour= "class magnitude")+
  geom_smooth(se = T, method = "loess", aes(group = class))+
  guides(color = guide_legend(override.aes = list(fill = NA)),
         linetype = guide_legend(override.aes = list(fill = NA))) +
  theme(panel.grid.major = element_blank(), # optional - remove gridlines
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        #legend.box.background = element_rect(colour = "red"),
        legend.key = element_rect(fill = "white", colour = "white"))+
  #scale_color_lancet()+
  scale_color_lancet(labels = c("217 trajectories (74.3%)", "75 trajectories (25.7%)")) 


dr4 <- read.csv("drs2_2020_2023_noq1.csv", sep=",") 

ggplot(dr4, aes(x=ano2, y=drs_coverage, colour = as.factor(class), group = comuna2)) +
  geom_point(size = .1) + 
  geom_line(size = .05) +
  geom_smooth(method='lm',formula=drs_coverage~ano2)+
  scale_x_continuous(breaks = c(0:12), labels=2011:2023)+
  xlab("Year") +
  ylab("Proportion of T2DM with an annual eye screening")+
  #theme_ipsum()+
  #labs(colour= "class magnitude")+
  geom_smooth(se = T, method = "loess", aes(group = class))+
  guides(color = guide_legend(override.aes = list(fill = NA)),
         linetype = guide_legend(override.aes = list(fill = NA))) +
  theme(panel.grid.major = element_blank(), # optional - remove gridlines
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        #legend.box.background = element_rect(colour = "red"),
        legend.key = element_rect(fill = "white", colour = "white"))+
  #scale_color_lancet()+
  scale_color_lancet(labels = c("245 trajectories (81.7%)", "55 trajectories (18.3%)")) 

```

```{r}
#| label: tbl-1
#| tbl-cap: "Numeric variables summary of diabetes coverage across two periods"
#| tbl-subcap: 
#| - "Descriptive statistics by municipalities (Municipality data average)"
#| - "Descriptive statistics of all raw data (Obsevation-based data)"
#| fig-align: center

#The mean of the means


# Read data
coverage_2011_2023_noq1 <- read.csv("coverage_2011_2023_noq1.csv", sep=",")


# Summarize data
data_summary1 <- modelsummary::datasummary_skim(coverage_2011_2023_noq1 %>% 
                                                  select("mean_dm_coverage", 
                                                         "mean_drs_coverage") %>% 
                                                  mutate_if(is.character, as.factor),
                                                type="numeric",
                                                histogram=T,
                                                fmt = "%.4f",
                                                output="gt") %>% 
  as.data.frame() 

data_summary2 <- modelsummary::datasummary_skim(coverage_2011_2023_noq1 %>% 
                                                  filter(ano2 %in% 1:9) %>% 
                                                  select("mean_dm_coverage", 
                                                         "mean_drs_coverage") %>% 
                                                  mutate_if(is.character, as.factor),
                                                type="numeric",
                                                histogram=T,
                                                fmt = "%.4f",
                                                output="gt") %>% 
                              as.data.frame()

data_summary3 <- rbind(data_summary1, data_summary2)

data_summary3 <- data_summary3 %>% 
  # Convert columns 4 to 8 to numeric
 mutate(across(4:8, ~ as.numeric(.))) %>% # Convert to numeric
  mutate(across(4:8, ~ round(. * 100, 1)))  %>% 
  dplyr::mutate(Metric = rep(c("DGCC", "DRSC"), 2)) %>% 
  select(-1, -"Missing Pct.")
 

# Print table using gt
data_summary3_gt <- data_summary3 %>% 
  dplyr::select(Metric, everything()) %>% 
  gt() %>%
  tab_options(
    table.font.size = 10,
    data_row.padding = px(1),
    table.border.top.color = "black",
    heading.border.bottom.color = "black",
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table_body.hlines.color = "white"
  ) %>% 
   tab_spanner(
    label = "Coverage Statistics (%)",
    columns = 3:7
    ) %>% 
   tab_spanner(
    label = "Municipality Level",
    columns = 1:7
    ) %>% 
  tab_row_group(
    group = "2011-2019",
    rows = 3:4
  ) %>%
  tab_row_group(
    group = "2011-2023",
    rows = 1:2
  )%>% 
 tab_source_note(source_note = md(" **Abbreviations:** DGCC - Diabetic glycemic control coverage; DRSC - Diabetic retinopathy screening coverage, ; ISDE - Index of Socioeconomic Development"))

data_summary3_gt



```

```{r}
#| label: tbl-2
#| tbl-cap: "Summary statistics of diabetes coverage and sociodemographics in Chilean municipalities across two periods"
#| fig-align: center


zone_summary <- coverage_2011_2023_noq1 %>%
  group_by(zona) %>%
  dplyr::summarise(
    N_comuna = n(),
    percent_comuna = N_comuna / nrow(coverage_2011_2023_noq1) * 100,
    longitudinal_mean_drs_coverage = mean(mean_drs_coverage, na.rm = TRUE),
    longitudinal_sd_drs_coverage = sd(mean_drs_coverage, na.rm = TRUE),
    longitudinal_mean_dm_coverage = mean(mean_dm_coverage, na.rm = TRUE),
    longitudinal_sd_dm_coverage = sd(mean_dm_coverage, na.rm = TRUE),
    mean_index_standardized = mean(index_standardized, na.rm = TRUE),
    sd_index_standardized = sd(index_standardized, na.rm = TRUE),
    mean_index= mean(index, na.rm = TRUE),
    sd_index = sd(index, na.rm = TRUE)) %>% 
  dplyr::rename(Demographic =zona)



classification_summary <- coverage_2011_2023_noq1 %>%
  group_by(urbanisation_classification) %>%
  dplyr::summarise(
    N_comuna = n(),
    percent_comuna = N_comuna / nrow(coverage_2011_2023_noq1) * 100,
    longitudinal_mean_drs_coverage = mean(mean_drs_coverage, na.rm = TRUE),
    longitudinal_sd_drs_coverage = sd(mean_drs_coverage, na.rm = TRUE),
    longitudinal_mean_dm_coverage = mean(mean_dm_coverage, na.rm = TRUE),
    longitudinal_sd_dm_coverage = sd(mean_dm_coverage, na.rm = TRUE),
    mean_index_standardized = mean(index_standardized, na.rm = TRUE),
    sd_index_standardized = sd(index_standardized, na.rm = TRUE),
    mean_index= mean(index, na.rm = TRUE),
    sd_index = sd(index, na.rm = TRUE)) %>% 
  dplyr::rename(Demographic = urbanisation_classification)



sociodemographics1 <- rbind(zone_summary, classification_summary)


zone_summary2 <- coverage_2011_2023_noq1 %>%
  filter(ano2 %in% 1:9) %>% 
  group_by(zona) %>%
    dplyr::summarise(
    N_comuna = n(),
    percent_comuna = N_comuna / nrow(coverage_2011_2023_noq1 %>% filter(ano2 %in% 1:9) ) * 100,
    longitudinal_mean_drs_coverage = mean(mean_drs_coverage, na.rm = TRUE),
    longitudinal_sd_drs_coverage = sd(mean_drs_coverage, na.rm = TRUE),
    longitudinal_mean_dm_coverage = mean(mean_dm_coverage, na.rm = TRUE),
    longitudinal_sd_dm_coverage = sd(mean_dm_coverage, na.rm = TRUE),
    mean_index_standardized = mean(index_standardized, na.rm = TRUE),
    sd_index_standardized = sd(index_standardized, na.rm = TRUE),
    mean_index= mean(index, na.rm = TRUE),
    sd_index = sd(index, na.rm = TRUE)) %>% 
  dplyr::rename(Demographic =zona)


classification_summary2 <- coverage_2011_2023_noq1 %>%
  filter(ano2 %in% 1:9) %>%
  group_by(urbanisation_classification) %>%
    dplyr::summarise(
    N_comuna = n(),
    percent_comuna = N_comuna / nrow(coverage_2011_2023_noq1 %>% filter(ano2 %in% 1:9) ) * 100,
    longitudinal_mean_drs_coverage = mean(mean_drs_coverage, na.rm = TRUE),
    longitudinal_sd_drs_coverage = sd(mean_drs_coverage, na.rm = TRUE),
    longitudinal_mean_dm_coverage = mean(mean_dm_coverage, na.rm = TRUE),
    longitudinal_sd_dm_coverage = sd(mean_dm_coverage, na.rm = TRUE),
    mean_index_standardized = mean(index_standardized, na.rm = TRUE),
    sd_index_standardized = sd(index_standardized, na.rm = TRUE),
    mean_index= mean(index, na.rm = TRUE),
    sd_index = sd(index, na.rm = TRUE)) %>% 
  dplyr::rename(Demographic = urbanisation_classification)


sociodemographics2 <- rbind(zone_summary2, classification_summary2)

sociodemographics <- rbind(sociodemographics2, sociodemographics1)

sociodemographics <- sociodemographics %>%  
  select(Demographic, N_comuna, percent_comuna, 
         longitudinal_mean_dm_coverage, longitudinal_sd_dm_coverage, 
         longitudinal_mean_drs_coverage, longitudinal_sd_drs_coverage,
         mean_index_standardized, sd_index_standardized) %>% 
  mutate(across(4:7, ~ . * 100)) %>% 
  mutate_if(is.numeric, round,2) %>% 
  mutate(Demographic = case_when(
    Demographic == "centro" ~ "Center",
    Demographic == "norte" ~ "North",
    Demographic == "sur" ~ "South",
    Demographic == "Mixta" ~ "Mixed",
    Demographic == "Rural" ~ "Rural",
    Demographic == "Urbana" ~ "Urban",
    TRUE ~ Demographic
  ))
 

sociodemographics %>% 
  gt::gt() %>%
  #tab_spanner(label = 'Model fit and diagnostic criteria', columns = 2:16) %>%
  tab_options(
    table.font.size = 10,
    data_row.padding = px(1),
    table.border.top.color = "black",
    heading.border.bottom.color = "black",
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table_body.hlines.color = "white") %>% 
  tab_row_group(group = "Zone (2011-2023)",
rows = c(7:9)
  ) %>%
  tab_row_group(group = "Urbanicity (2011-2023)",
rows = c(10:12)
  ) %>%
  tab_row_group(
    group = "Zone (2011-2019)",
rows = c(1:3)
  )  %>%
  tab_row_group(
    group = "Urbanicity (2011-2019)",
rows = c(4:6)
  ) %>% 
  cols_label(N_comuna = "N",
             percent_comuna = "%",
             longitudinal_mean_drs_coverage = "Mean",
             longitudinal_sd_drs_coverage = "SD",
             longitudinal_mean_dm_coverage = "Mean",
             longitudinal_sd_dm_coverage = "SD",
             mean_index_standardized = "Mean",
             sd_index_standardized = "SD") %>%
  tab_spanner(
    label = "Standarized ISDE",
    columns = c("mean_index_standardized", "sd_index_standardized")
  ) %>% 
  tab_spanner(
    label = "DGCC (%)",
    columns = c("longitudinal_mean_dm_coverage", "longitudinal_sd_dm_coverage")
  ) %>% 
  tab_spanner(
    label = "DRSC (%)",
    columns = c("longitudinal_mean_drs_coverage", "longitudinal_sd_drs_coverage")
  ) %>% 
  tab_spanner(
    label = "Frequency",
    columns = c("N_comuna", "percent_comuna")
  ) %>% 
   tab_source_note(source_note = md(" **Abbreviations:** DGCC - Diabetic glycemic control coverage; DRSC - Diabetic retinopathy screening coverage; ISDE - Index of Socioeconomic Development; N - number of occurrences or observations within each category."))




```

```{r}
#| label: tbl-3
#| tbl-cap: "Latent class trajectory mixture models for diabetic glycemic control and diabetic retinopathy screening coverage across 300 (2011-2023) and 292 (2011-2019) municipalities"
#| fig-align: left

#DGCC 2011-2023
model_adequacy <- read.csv("model_adequacy_table.csv", sep=",") 


mfc1 <- model_adequacy %>% 
  filter(str_detect(Model, "drsc_model_2011_2023")) %>% 
  dplyr::rename(
    "LCMM class" = G, 
    "Maximum log-likelihood" = loglik,
    "Nº of parameters" = npm,
    "Smallest class count (n)" = Smallest_Class_Count,
    "Smallest class size (%)" = Smallest_Class_Size_Percentage)

mfc2 <- model_adequacy %>% 
  filter(str_detect(Model, "drsc_model_2011_2019")) %>% 
  dplyr::rename(
    "LCMM class" = G, 
    "Maximum log-likelihood" = loglik,
    "Nº of parameters" = npm,
    "Smallest class count (n)" = Smallest_Class_Count,
    "Smallest class size (%)" = Smallest_Class_Size_Percentage)

mfc3 <- model_adequacy %>% 
  filter(str_detect(Model, "dgcc_model_2011_2023")) %>% 
  dplyr::rename(
    "LCMM class" = G, 
    "Maximum log-likelihood" = loglik,
    "Nº of parameters" = npm,
    "Smallest class count (n)" = Smallest_Class_Count,
    "Smallest class size (%)" = Smallest_Class_Size_Percentage)

mfc4 <- model_adequacy %>% 
  filter(str_detect(Model, "dgcc_model_2011_2019")) %>% 
  dplyr::rename(
    "LCMM class" = G, 
    "Maximum log-likelihood" = loglik,
    "Nº of parameters" = npm,
    "Smallest class count (n)" = Smallest_Class_Count,
    "Smallest class size (%)" = Smallest_Class_Size_Percentage)




table_dgcc <- rbind(mfc1, mfc2)

table_drsc <- rbind(mfc3, mfc4)


class_solutions <- rbind(table_dgcc, table_drsc)

options(scipen = 999)  # reduce scientific notation globally


class_solutions <- class_solutions %>% 
  filter(str_detect(Model, "linear_nre_homocedastic_")) %>% 
  select(Model, `LCMM class`,"Nº of parameters", BIC) %>% 
  mutate(BIC = BIC %>% round(1)) %>% 
  select(-Model) %>% 
gt::gt() %>%
   tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      columns = BIC,
      rows = BIC == min(BIC)
    )
  ) %>% 
#tab_spanner(label = 'Model fit and diagnostic criteria', columns = 2:16) %>%
tab_options(
table.font.size = 10,
data_row.padding = px(1),
table.border.top.color = "black",
heading.border.bottom.color = "black",
row_group.border.top.color = "black",
row_group.border.bottom.color = "white",
table.border.bottom.color = "white",
column_labels.border.top.color = "black",
column_labels.border.bottom.color = "black",
table_body.border.bottom.color = "black",
table_body.hlines.color = "white")%>%
   tab_row_group(
    group = "Diabetic retinopathy screening coverage 2011-2019 
(n= 300 municipalities)",
    rows = 15:21
  ) %>% 
  tab_row_group(
    group = "Diabetic glycemic control coverage 2011-2019 
(n= 292 municipalities)",
    rows =22:28
  ) %>% 
   tab_row_group(
    group = "Diabetic retinopathy screening coverage 2011-2023 
(n= 292 municipalities)",
    rows = 1:7
 ) %>% 
   tab_row_group(
    group = "Diabetic glycemic control coverage 2011-2023
(n= 300 municipalities)",
    rows = 8:14
 ) %>% 
   tab_source_note(source_note = md(" **Abbreviations:** LCMM - Latent Class Mixture Model; MLL – Maximum Log-likelihood; AIC- Akaike Information Criteria; BIC - Bayesian Information Criteria; SABIC - Sample-Adjusted Bayesian Information Criteria; CAIC -  Consistent Akaike Information Criterion; AWE- Approximate Weight of Evidence Criterion; SIC - Schwarz Information Criterion; BF - Bayes Factor; cmP- correct model probability; SCC - Smallest Class Count; SCS - Smallest Class Size; APPA: Average Class Posterior Probability; MMV: Mistmatch Value; OCC: Odds od Correct Classification; VLMRLRT: Vuong–Lo–Mendell–Rubin Likelihood Radio Test"))
 
class_solutions
```

```{r}
#| label: tbl-4
#| tbl-cap: "Diabetic glycemic control coverage trajectory class characteristics 2011-2023" 
#| fig-align: center

df_dm_2011_2023 <- read.csv("df_dm_2011_2023.csv", sep=",") 

df_dm_2011_2023 <- df_dm_2011_2023 %>% 
  select(-1) 

# Round numeric p-values
df_dm_2011_2023 <- df_dm_2011_2023 %>%
  mutate(across(contains("P"), ~ifelse(!is.na(.x) & .x != "<0.001", sprintf("%.3f", as.numeric(.x)), .x)))

# Multiply columns containing "B" by 100 and round to two decimals
df_dm_2011_2023 <- df_dm_2011_2023 %>%
  dplyr::mutate(across(contains("B"), ~ifelse(!is.na(.x), .x * 100, .x))) %>%
  dplyr::mutate(across(contains("SE"), ~ifelse(!is.na(.x), .x * 100, .x))) %>% 
  dplyr::mutate(across(contains(c("B", "SE")), ~ifelse(!is.na(.x), round(.x, 2), .x)))

# Replace NAs with empty spaces
df_dm_2011_2023[is.na(df_dm_2011_2023)] <- ""

# Rename columns for display
colnames(df_dm_2011_2023) <- c(
  "Metric",
  "B1", "SE1", "p-value1",
  "B2", "SE2", "p-value2",
  "B3", "SE3", "p-value3",
  "B4", "SE4", "p-value4",
  "B5", "SE5", "p-value5",
  "B6", "SE6", "p-value6"
)



# Create the gt table with tab_spanner
gt::gt(df_dm_2011_2023) %>%
  cols_label(
    B1 = "B", SE1 = "SE", `p-value1` = "p-value",
    B2 = "B", SE2 = "SE", `p-value2` = "p-value",
    B3 = "B", SE3 = "SE", `p-value3` = "p-value",
    B4 = "B", SE4 = "SE", `p-value4` = "p-value",
    B5 = "B", SE5 = "SE", `p-value5` = "p-value",
    B6 = "B", SE6 = "SE", `p-value6` = "p-value"
  ) %>%
  tab_spanner(
    label = "Six-class Model",
    columns = c("B6", "SE6", "p-value6")
  ) %>% 
  tab_spanner(
    label = "Five-class Model",
    columns = c("B5", "SE5", "p-value5")
  ) %>% 
  tab_spanner(
    label = "Four-class Model",
    columns = c("B4", "SE4", "p-value4")
  ) %>% 
  tab_spanner(
    label = "Three-class Model",
    columns = c("B3", "SE3", "p-value3")
  ) %>% 
  tab_spanner(
    label = "Two-class Model",
    columns = c("B2", "SE2", "p-value2")
  ) %>% 
  tab_spanner(
    label = "One-class Model",
    columns = c("B1", "SE1", "p-value1")
  ) %>% 
  tab_options(
    table.font.size = 10,
    data_row.padding = px(1),
    table.border.top.color = "black",
    heading.border.bottom.color = "black",
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table_body.hlines.color = "white"
  ) %>%
  tab_row_group(
    group = "Class 6",
    rows = 11:12
  ) %>%
  tab_row_group(
    group = "Class 5",
    rows = 9:10
  ) %>%
  tab_row_group(
    group = "Class 4",
    rows = 7:8
  ) %>%
  tab_row_group(
    group = "Class 3",
    rows = 5:6
  ) %>%
  tab_row_group(
    group = "Class 2",
    rows = 3:4
  ) %>%
  tab_row_group(
    group = "Class 1",
    rows = 1:2
  ) %>% 
  tab_source_note(source_note = md(" **Abbreviations / notes :** B - is the intercept (or constant term), which represents the value of 𝑌 Y when all predictors are zero; SE - Standard Error. Note: p-value - Wald test measures the significance of each coefficient in the regression model"))



```

```{r}
#| label: tbl-5
#| tbl-cap: "Diabetic glycemic control coverage trajectory class characteristics 2011-2019" 
#| fig-align: center


df_dm_2011_2019 <- read.csv("df_dm_2011_2019.csv", sep=",") 

df_dm_2011_2019 <- df_dm_2011_2019 %>% 
  select(-1) 

# Round numeric p-values
df_dm_2011_2019 <- df_dm_2011_2019 %>%
  mutate(across(contains("P"), ~ifelse(!is.na(.x) & .x != "<0.001", sprintf("%.3f", as.numeric(.x)), .x)))

# Multiply columns containing "B" by 100 and round to two decimals
df_dm_2011_2019 <- df_dm_2011_2019 %>%
  dplyr::mutate(across(contains("B"), ~ifelse(!is.na(.x), .x * 100, .x))) %>%
  dplyr::mutate(across(contains("SE"), ~ifelse(!is.na(.x), .x * 100, .x))) %>% 
  dplyr::mutate(across(contains(c("B", "SE")), ~ifelse(!is.na(.x), round(.x, 2), .x)))

# Replace NAs with empty spaces
df_dm_2011_2019[is.na(df_dm_2011_2019)] <- ""

# Rename columns for display
colnames(df_dm_2011_2019) <- c(
  "Metric",
  "B1", "SE1", "p-value1",
  "B2", "SE2", "p-value2",
  "B3", "SE3", "p-value3",
  "B4", "SE4", "p-value4",
  "B5", "SE5", "p-value5",
  "B6", "SE6", "p-value6"
)



# Create the gt table with tab_spanner
gt::gt(df_dm_2011_2019) %>%
  cols_label(
    B1 = "B", SE1 = "SE", `p-value1` = "p-value",
    B2 = "B", SE2 = "SE", `p-value2` = "p-value",
    B3 = "B", SE3 = "SE", `p-value3` = "p-value",
    B4 = "B", SE4 = "SE", `p-value4` = "p-value",
    B5 = "B", SE5 = "SE", `p-value5` = "p-value",
    B6 = "B", SE6 = "SE", `p-value6` = "p-value"
  ) %>%
  tab_spanner(
    label = "Six-class Model",
    columns = c("B6", "SE6", "p-value6")
  ) %>% 
  tab_spanner(
    label = "Five-class Model",
    columns = c("B5", "SE5", "p-value5")
  ) %>% 
  tab_spanner(
    label = "Four-class Model",
    columns = c("B4", "SE4", "p-value4")
  ) %>% 
  tab_spanner(
    label = "Three-class Model",
    columns = c("B3", "SE3", "p-value3")
  ) %>% 
  tab_spanner(
    label = "Two-class Model",
    columns = c("B2", "SE2", "p-value2")
  ) %>% 
  tab_spanner(
    label = "One-class Model",
    columns = c("B1", "SE1", "p-value1")
  ) %>% 
  tab_options(
    table.font.size = 10,
    data_row.padding = px(1),
    table.border.top.color = "black",
    heading.border.bottom.color = "black",
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table_body.hlines.color = "white"
  ) %>%
  tab_row_group(
    group = "Class 6",
    rows = 11:12
  ) %>%
  tab_row_group(
    group = "Class 5",
    rows = 9:10
  ) %>%
  tab_row_group(
    group = "Class 4",
    rows = 7:8
  ) %>%
  tab_row_group(
    group = "Class 3",
    rows = 5:6
  ) %>%
  tab_row_group(
    group = "Class 2",
    rows = 3:4
  ) %>%
  tab_row_group(
    group = "Class 1",
    rows = 1:2
  ) %>% 
  tab_source_note(source_note = md(" **Abbreviations / notes :** B - is the intercept (or constant term), which represents the value of 𝑌 Y when all predictors are zero; SE - Standard Error. Note: p-value - Wald test measures the significance of each coefficient in the regression model"))







```

```{r}
#| label: tbl-6
#| tbl-cap: "Diabetic retinopathy screening coverage trajectory class characteristics 2011-2023" 
#| fig-align: center



df_drs_2011_2023 <- read.csv("df_drs_2011_2023.csv", sep=",") 

df_drs_2011_2023 <- df_drs_2011_2023 %>% 
  select(-1) 

# Round numeric p-values
df_drs_2011_2023 <- df_drs_2011_2023 %>%
  mutate(across(contains("P"), ~ifelse(!is.na(.x) & .x != "<0.001", sprintf("%.3f", as.numeric(.x)), .x)))

# Multiply columns containing "B" by 100 and round to two decimals
df_drs_2011_2023 <- df_drs_2011_2023 %>%
  dplyr::mutate(across(contains("B"), ~ifelse(!is.na(.x), .x * 100, .x))) %>%
  dplyr::mutate(across(contains("SE"), ~ifelse(!is.na(.x), .x * 100, .x))) %>% 
  dplyr::mutate(across(contains(c("B", "SE")), ~ifelse(!is.na(.x), round(.x, 2), .x)))

# Replace NAs with empty spaces
df_drs_2011_2023[is.na(df_drs_2011_2023)] <- ""

# Rename columns for display
colnames(df_drs_2011_2023) <- c(
  "Metric",
  "B1", "SE1", "p-value1",
  "B2", "SE2", "p-value2",
  "B3", "SE3", "p-value3",
  "B4", "SE4", "p-value4",
  "B5", "SE5", "p-value5",
  "B6", "SE6", "p-value6"
)



# Create the gt table with tab_spanner
gt::gt(df_drs_2011_2023) %>%
  cols_label(
    B1 = "B", SE1 = "SE", `p-value1` = "p-value",
    B2 = "B", SE2 = "SE", `p-value2` = "p-value",
    B3 = "B", SE3 = "SE", `p-value3` = "p-value",
    B4 = "B", SE4 = "SE", `p-value4` = "p-value",
    B5 = "B", SE5 = "SE", `p-value5` = "p-value",
    B6 = "B", SE6 = "SE", `p-value6` = "p-value"
  ) %>%
  tab_spanner(
    label = "Six-class Model",
    columns = c("B6", "SE6", "p-value6")
  ) %>% 
  tab_spanner(
    label = "Five-class Model",
    columns = c("B5", "SE5", "p-value5")
  ) %>% 
  tab_spanner(
    label = "Four-class Model",
    columns = c("B4", "SE4", "p-value4")
  ) %>% 
  tab_spanner(
    label = "Three-class Model",
    columns = c("B3", "SE3", "p-value3")
  ) %>% 
  tab_spanner(
    label = "Two-class Model",
    columns = c("B2", "SE2", "p-value2")
  ) %>% 
  tab_spanner(
    label = "One-class Model",
    columns = c("B1", "SE1", "p-value1")
  ) %>% 
  tab_options(
    table.font.size = 10,
    data_row.padding = px(1),
    table.border.top.color = "black",
    heading.border.bottom.color = "black",
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table_body.hlines.color = "white"
  ) %>%
  tab_row_group(
    group = "Class 6",
    rows = 11:12
  ) %>%
  tab_row_group(
    group = "Class 5",
    rows = 9:10
  ) %>%
  tab_row_group(
    group = "Class 4",
    rows = 7:8
  ) %>%
  tab_row_group(
    group = "Class 3",
    rows = 5:6
  ) %>%
  tab_row_group(
    group = "Class 2",
    rows = 3:4
  ) %>%
  tab_row_group(
    group = "Class 1",
    rows = 1:2
  ) %>% 
  tab_source_note(source_note = md(" **Abbreviations / notes :** B - is the intercept (or constant term), which represents the value of 𝑌 Y when all predictors are zero; SE - Standard Error. Note: p-value - Wald test measures the significance of each coefficient in the regression model"))
```

```{r}
#| label: tbl-7
#| tbl-cap: "Diabetic retinopathy screening coverage trajectory class characteristics 2011-2019" 
#| fig-align: center


df_drs_2011_2019 <- read.csv("df_drs_2011_2019.csv", sep=",") 

df_drs_2011_2019 <- df_drs_2011_2019 %>% 
  select(-1) 

# Round numeric p-values
df_drs_2011_2019 <- df_drs_2011_2019 %>%
  mutate(across(contains("P"), ~ifelse(!is.na(.x) & .x != "<0.001", sprintf("%.3f", as.numeric(.x)), .x)))

# Multiply columns containing "B" by 100 and round to two decimals
df_drs_2011_2019 <- df_drs_2011_2019 %>%
  dplyr::mutate(across(contains("B"), ~ifelse(!is.na(.x), .x * 100, .x))) %>%
  dplyr::mutate(across(contains("SE"), ~ifelse(!is.na(.x), .x * 100, .x))) %>% 
  dplyr::mutate(across(contains(c("B", "SE")), ~ifelse(!is.na(.x), round(.x, 2), .x)))

# Replace NAs with empty spaces
df_drs_2011_2019[is.na(df_drs_2011_2019)] <- ""

# Rename columns for display
colnames(df_drs_2011_2019) <- c(
  "Metric",
  "B1", "SE1", "p-value1",
  "B2", "SE2", "p-value2",
  "B3", "SE3", "p-value3",
  "B4", "SE4", "p-value4",
  "B5", "SE5", "p-value5",
  "B6", "SE6", "p-value6"
)



# Create the gt table with tab_spanner
gt::gt(df_drs_2011_2019) %>%
  cols_label(
    B1 = "B", SE1 = "SE", `p-value1` = "p-value",
    B2 = "B", SE2 = "SE", `p-value2` = "p-value",
    B3 = "B", SE3 = "SE", `p-value3` = "p-value",
    B4 = "B", SE4 = "SE", `p-value4` = "p-value",
    B5 = "B", SE5 = "SE", `p-value5` = "p-value",
    B6 = "B", SE6 = "SE", `p-value6` = "p-value"
  ) %>%
  tab_spanner(
    label = "Six-class Model",
    columns = c("B6", "SE6", "p-value6")
  ) %>% 
  tab_spanner(
    label = "Five-class Model",
    columns = c("B5", "SE5", "p-value5")
  ) %>% 
  tab_spanner(
    label = "Four-class Model",
    columns = c("B4", "SE4", "p-value4")
  ) %>% 
  tab_spanner(
    label = "Three-class Model",
    columns = c("B3", "SE3", "p-value3")
  ) %>% 
  tab_spanner(
    label = "Two-class Model",
    columns = c("B2", "SE2", "p-value2")
  ) %>% 
  tab_spanner(
    label = "One-class Model",
    columns = c("B1", "SE1", "p-value1")
  ) %>% 
  tab_options(
    table.font.size = 10,
    data_row.padding = px(1),
    table.border.top.color = "black",
    heading.border.bottom.color = "black",
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table_body.hlines.color = "white"
  ) %>%
  tab_row_group(
    group = "Class 6",
    rows = 11:12
  ) %>%
  tab_row_group(
    group = "Class 5",
    rows = 9:10
  ) %>%
  tab_row_group(
    group = "Class 4",
    rows = 7:8
  ) %>%
  tab_row_group(
    group = "Class 3",
    rows = 5:6
  ) %>%
  tab_row_group(
    group = "Class 2",
    rows = 3:4
  ) %>%
  tab_row_group(
    group = "Class 1",
    rows = 1:2
  ) %>% 
  tab_source_note(source_note = md(" **Abbreviations / notes :** B - is the intercept (or constant term), which represents the value of 𝑌 Y when all predictors are zero; SE - Standard Error. Note: p-value - Wald test measures the significance of each coefficient in the regression model"))




```

```{r}
#| label: tbl-8
#| tbl-cap: "Diabetic retinopathy screening coverage trajectory class characteristics 2020-2023" 
#| fig-align: center



df_drs_2020_2023 <- read.csv("df_drs_2020_2023.csv", sep=",") 

df_drs_2020_2023 <- df_drs_2020_2023 %>% 
  select(-1) 

# Round numeric p-values
df_drs_2020_2023 <- df_drs_2020_2023 %>%
  mutate(across(contains("P"), ~ifelse(!is.na(.x) & .x != "<0.001", sprintf("%.3f", as.numeric(.x)), .x)))

# Multiply columns containing "B" by 100 and round to two decimals
df_drs_2020_2023 <- df_drs_2020_2023 %>%
  dplyr::mutate(across(contains("B"), ~ifelse(!is.na(.x), .x * 100, .x))) %>%
  dplyr::mutate(across(contains("SE"), ~ifelse(!is.na(.x), .x * 100, .x))) %>% 
  dplyr::mutate(across(contains(c("B", "SE")), ~ifelse(!is.na(.x), round(.x, 2), .x)))

# Replace NAs with empty spaces
df_drs_2020_2023[is.na(df_drs_2020_2023)] <- ""

# Rename columns for display
colnames(df_drs_2020_2023) <- c(
  "Metric",
  "B1", "SE1", "p-value1",
  "B2", "SE2", "p-value2",
  "B3", "SE3", "p-value3",
  "B4", "SE4", "p-value4",
  "B5", "SE5", "p-value5",
  "B6", "SE6", "p-value6"
)



# Create the gt table with tab_spanner
gt::gt(df_drs_2020_2023) %>%
  cols_label(
    B1 = "B", SE1 = "SE", `p-value1` = "p-value",
    B2 = "B", SE2 = "SE", `p-value2` = "p-value",
    B3 = "B", SE3 = "SE", `p-value3` = "p-value",
    B4 = "B", SE4 = "SE", `p-value4` = "p-value",
    B5 = "B", SE5 = "SE", `p-value5` = "p-value",
    B6 = "B", SE6 = "SE", `p-value6` = "p-value"
  ) %>%
  tab_spanner(
    label = "Six-class Model",
    columns = c("B6", "SE6", "p-value6")
  ) %>% 
  tab_spanner(
    label = "Five-class Model",
    columns = c("B5", "SE5", "p-value5")
  ) %>% 
  tab_spanner(
    label = "Four-class Model",
    columns = c("B4", "SE4", "p-value4")
  ) %>% 
  tab_spanner(
    label = "Three-class Model",
    columns = c("B3", "SE3", "p-value3")
  ) %>% 
  tab_spanner(
    label = "Two-class Model",
    columns = c("B2", "SE2", "p-value2")
  ) %>% 
  tab_spanner(
    label = "One-class Model",
    columns = c("B1", "SE1", "p-value1")
  ) %>% 
  tab_options(
    table.font.size = 10,
    data_row.padding = px(1),
    table.border.top.color = "black",
    heading.border.bottom.color = "black",
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table_body.hlines.color = "white"
  ) %>%
  tab_row_group(
    group = "Class 6",
    rows = 11:12
  ) %>%
  tab_row_group(
    group = "Class 5",
    rows = 9:10
  ) %>%
  tab_row_group(
    group = "Class 4",
    rows = 7:8
  ) %>%
  tab_row_group(
    group = "Class 3",
    rows = 5:6
  ) %>%
  tab_row_group(
    group = "Class 2",
    rows = 3:4
  ) %>%
  tab_row_group(
    group = "Class 1",
    rows = 1:2
  ) %>% 
  tab_source_note(source_note = md(" **Abbreviations / notes :** B - is the intercept (or constant term), which represents the value of 𝑌 Y when all predictors are zero; SE - Standard Error. Note: p-value - Wald test measures the significance of each coefficient in the regression model"))


```

```{r}
#| label: tbl-9
#| tbl-cap: "Logistic regression 2011-2019"
#| fig-align: center
#| results: "asis"


# Model 5: Single predictor (index_standardized)
model5 <- glm(class_membership ~ index_standardized, family = "binomial", data = d2)
# Model 5a: Single predictor (classification)
model5a <- glm(class_membership ~ classification, family = "binomial", data = d2)
# Model 5b: Single predictor (zona)
model5b <- glm(class_membership ~ zona, family = "binomial", data = d2)
# Model 5c: All predictors
model5c <- glm(class_membership ~ index_standardized + classification + zona, family = "binomial", data = d2)
# Model 6: Two predictors (index_standardized + classification)
model6 <- glm(class_membership ~ index_standardized + classification, family = "binomial", data = d2)
# Model 6a: Two predictors (index_standardized + zona)
model6a <- glm(class_membership ~ index_standardized + zona, family = "binomial", data = d2)
# Model 6b: Two predictors (classification + zona)
model6b <- glm(class_membership ~ classification + zona, family = "binomial", data = d2)



models2 <- list("M1" = model5, 
               "M2" = model5a, 
               "M3" = model5b,
               "M4" = model6, 
               "M5" = model6a,
               "M6" = model6b,
               "M7" = model5c)
  
# additionally we want to change the font, font size and spacing
modelsummary(models2,
output = 'gt',
coef_map = c("index_standardized" = "Deprivation (Standarized ISDE score)",
             "classificationRural" = "Urbanicity (Rural)",
             "classificationUrbana" = "Urbanicity (Urban)",
            "zonanorte" = "Zone (North)",
            "zonasur" = "Zone (South)"),
#stars = T,
#estimate = "{estimate}{stars}",
statistic = c("SE = {std.error}", 
                "p = {p.value}{stars}"),
notes = "Notes: + p < 0.1, * p < 0.05, ** p < 0.01, *** p < 0.001") %>%
tab_spanner(label = 'Dependent variable: DRSC class', columns = 2:8) %>%
tab_options(
table.font.size = 10,
data_row.padding = px(1),
table.border.top.color = "white",
heading.border.bottom.color = "black",
row_group.border.top.color = "black",
row_group.border.bottom.color = "white",
table.border.bottom.color = "white",
column_labels.border.top.color = "black",
column_labels.border.bottom.color = "black",
table_body.border.bottom.color = "black",
table_body.hlines.color = "white"
) %>% 
  tab_row_group(
    group = "",
    rows = 1:15
  )




```

```{r}
#| label: tbl-10
#| tbl-cap: "Logistic regression 2011-2023"
#| fig-align: left
#| results: "asis"


# Model 1: Single predictor (index_standardized)
model1 <- glm(class_membership ~ index_standardized, family = "binomial", data = d1)
# Model 1a: Single predictor (classification)
model1a <- glm(class_membership ~ classification, family = "binomial", data = d1)
# Model 1b: Single predictor (zona)
model1b <- glm(class_membership ~ zona, family = "binomial", data = d1)
# Model 1c: All predictors
model1c <- glm(class_membership ~ index_standardized + classification + zona, family = "binomial", data = d1)
# Model 2: Two predictors (index_standardized + classification)
model2 <- glm(class_membership ~ index_standardized + classification, family = "binomial", data = d1)
# Model 2a: Two predictors (index_standardized + zona)
model2a <- glm(class_membership ~ index_standardized + zona, family = "binomial", data = d1)
# Model 2b: Two predictors (classification + zona)
model2b <- glm(class_membership ~ classification + zona, family = "binomial", data = d1)

models <- list("M1" = model1, 
               "M2" = model1a, 
               "M3" = model1b,
               "M4" = model2, 
               "M5" = model2a,
               "M6" = model2b,
               "M7" = model1c)
  
# additionally we want to change the font, font size and spacing
modelsummary(models,
output = 'gt',
coef_map = c("index_standardized" = "Deprivation (Standarized ISDE score)",
             "classificationRural" = "Urbanicity (Rural)",
             "classificationUrbana" = "Urbanicity (Urban)",
            "zonanorte" = "Zona (Norte)",
            "zonasur" = "Zona (Sur)"),
#stars = T,
#estimate = "{estimate}{stars}",
statistic = c("SE = {std.error}", 
                "p = {p.value}{stars}"),
notes = "Notes: + p < 0.1, * p < 0.05, ** p < 0.01, *** p < 0.001") %>%
tab_spanner(label = 'Dependent variable: DRSC class', columns = 2:8) %>%
tab_options(
table.font.size = 10,
data_row.padding = px(1),
table.border.top.color = "white",
heading.border.bottom.color = "black",
row_group.border.top.color = "black",
row_group.border.bottom.color = "white",
table.border.bottom.color = "white",
column_labels.border.top.color = "black",
column_labels.border.bottom.color = "black",
table_body.border.bottom.color = "black",
table_body.hlines.color = "white"
) %>% 
  tab_row_group(
    group = "",
    rows = 1:15
  )





```

```{r}
#| label: tbl-11
#| tbl-cap: "Reporting of latent growth modeling approach using the GRoLTS checklist"
#| fig-align: left


grolts <- read.csv("Grolts_report1.csv", sep=",") 


grolts %>% 
  dplyr::rename(Items = X) %>% 
  gt::gt() %>%
  #tab_spanner(label = 'Model fit and diagnostic criteria', columns = 2:16) %>%
  tab_options(
    table.font.size = 10,
    data_row.padding = px(1),
    table.border.top.color = "black",
    heading.border.bottom.color = "black",
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table_body.hlines.color = "white") 

```

```{r}
#| label: tbl-12
#| tbl-cap: "Framework of eight steps to construct a latent class trajectory model"
#| fig-align: left

tibble::tibble(
  Step = 1:8,
  `Step description` = c(
    "Scope model by provisionally selecting a plausible number of classes based on available literature and the structure based on plausible clinical patterns.",
    "Refine the model from step 1 to confirm the optimal number of classes, typically testing K=1–7 classes.",
    "Refine optimal model structure from fixed through to unrestricted random effects of the model using the favoured K derived in step 2.",
    "Run model adequacy assessments as described in online supplementary table S3 including posterior probability of assignments (APPA), odds of correct classification (OCC) and relative entropy.",
    "Investigate graphical presentation",
    "Run additional tools to assess discrimination including Degrees of separation (DoS) and Elsensohn’s envelope of residuals",
    "Assess for clinical characterisation and plausibility.",
    "Conduct sensitivity analyses, for example, testing models without complete data at all time points."
  ),
  `Criteria for selection` = c(
    "Examine linearity of the shape of standardised residual plots for each of the classes in a model with no random effects.",
    "Lowest Bayesian information criteria value.",
    "",
    "APPA: average of maximum probabilities should be greater than 70% for all classes.\nOCC values greater than 5.0.\nRelative entropy values greater than 0.5.",
    "Plot mean trajectories across time for each class in a single graph.\nPlot mean trajectories with 95% predictive intervals for each class (one class per graph).\nPlot individual class ‘spaghetti plots’ across time for a random sample.",
    "DoS greater than zero.\nEnvelope of residuals is assessed in plots by observing clear separations between classes.",
    "Tabulation of characteristics by latent classes. Are the trajectory patterns clinically meaningful? Perhaps, consider classes with a minimum percentage of the population.\nAre the trajectory patterns clinically plausible?\nConcordance of class characteristics with those for other well-established variables.",
    "General assessment of patterns of trajectories compared with main model."
  )
) %>%
  gt::gt() %>%
  #tab_spanner(label = 'Model fit and diagnostic criteria', columns = 2:16) %>%
  tab_options(
    table.font.size = 10,
    data_row.padding = px(1),
    table.border.top.color = "black",
    heading.border.bottom.color = "black",
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "black",
    column_labels.border.bottom.color = "black",
    table_body.border.bottom.color = "black",
    table_body.hlines.color = "white") 
```

📊 Example Summary Table (based on your plots): Class Residual Shape Pattern Max Drift (approx.) Polynomial Suggestion 1 Slight upward trend, mild curve \~0.3 Linear or quadratic 2 Mild U-shape \~0.3–0.4 Quadratic 3 Sine-like oscillation \~0.4–0.5 Cubic or spline 4 U-shape + slight drift \~0.4 Quadratic or cubic 5 Slight wave \~0.3 Cubic 6 Mild upward then downward \~0.3–0.4 Cubic 7 Flat-ish \~0.2 Linear might suffice

```{r}


residual_structure_drift_df <- data.frame(
  Class = c("Class 1", "Class 2", "Class 3", "Class 4", "Class 5", "Class 6", "Class 7"),
  Residual_Shape_Pattern = c(
    "Slight upward trend, mild curve",
    "Mild U-shape",
    "Sine-like oscillation",
    "U-shape + slight drift",
    "Slight wave",
    "Mild upward then downward",
    "Flat-ish"
  ),
  Max_Drift_Approx = c("~0.3", "~0.3–0.4", "~0.4–0.5", "~0.4", "~0.3", "~0.3–0.4", "~0.2"),
  Polynomial_Suggestion = c(
    "Linear or quadratic",
    "Quadratic",
    "Cubic or spline",
    "Quadratic or cubic",
    "Cubic",
    "Cubic",
    "Linear might suffice"
  ),
  stringsAsFactors = FALSE
)


```

Class Residual pattern Suggested structure Class 1 Slight downward curve early on, then flattens Mild quadratic trend Class 2 Slight upward drift, flattens in middle, then up again Quadratic (or light cubic) Class 3 Mild S-shape, mostly constant Possibly cubic (but marginal) Class 4 Linear central trend with small tail deviation Linear or mild quadratic Class 5 Light wave (S-pattern), more visible curvature Cubic trend likely helpful Class 6 Systematic wave (rise–dip–rise) even if small Cubic or higher polynomial suggested Class 7 Relatively flat, minor wiggle Possibly linear–quadratic hybrid (but low concern)
